---
title: "Homework 7"
author: "Ansel George"
output: pdf_document
---

```{r}
library(ggplot2)

set.seed(10)
```


# Poisson Processes

## Haldane's Model of Recombination

*Background: J.B.S. Haldane ([Wikipedia
page](http://en.wikipedia.org/wiki/J._B._S._Haldane)) modeled the rate of
crossovers as a Poisson process along a chromosome. As the number of crossovers
expected in a given genomic tract can vary depending on where you are in the
genome, Haldane used a distance measure referred to as genetic distance.
Genetic distance is measured in units of Morgans (M), where a Morgan is defined
such that there is one crossover event expected per Morgan (or 0.01 events per
centiMorgan, cM). Note that Morgans may correspond to different physical
distances depending on where you are in the genome.*

*Answer the following questions:*

### 1.
*For a chromosome of length $t$ centiMorgans (cM), what is the mean and variance
of the number of crossover events as a function of $t$?*

For the sake of simplicity, the units for $t$ are left in Morgans.

\begin{align}
P(t) &= \frac{(\lambda t)^n}{n!}e^{-\lambda t} \\
E[T] &= \sum_{i=0}^\infty n \frac{(\lambda t)^n}{n!}e^{-\lambda t} \\
  &= \sum_{i=0}^\infty \frac{(\lambda t)^n}{(n-1)!}e^{-\lambda t} \\
  &= \lambda t \sum_{i=0}^\infty \frac{(\lambda t)^{n-1}}{(n-1)!}e^{-\lambda t} \\ 
  &= \lambda t \sum_{k=1}^\infty \frac{(\lambda t)^{n-1}}{(n-1)!}e^{-\lambda t}, \textrm{ where } k=i+1 \\
  &= \lambda t \sum_{k=0}^\infty \frac{(\lambda t)^{n}}{(n)!}e^{-\lambda t} \\
  &= \lambda t \\
  &= \lambda t \rightarrow \frac{\lambda t}{100} \textrm{for t in units of cM}
\end{align}

\begin{align}
Var[P(t)] &= E[T^2] - (E[T])^2 \\
  &= E[T^2] - (\lambda t)^2 \\
  &= \sum_{i=0}^\infty n^2 \frac{(\lambda t)^n}{n!}e^{-\lambda t}  - (\lambda t)^2 \\
  &= \sum_{i=0}^\infty \frac{n (\lambda t)^n}{(n-1)!}e^{-\lambda t}  - (\lambda t)^2 \\
  &= \lambda t \sum_{i=0}^\infty \frac{n (\lambda t)^{n-1}}{(n-1)!}e^{-\lambda t}  - (\lambda t)^2 \\ 
  &= \lambda t \sum_{i=0}^\infty \frac{(n-1+1)(\lambda t)^{n-1}}{(n-1)!}e^{-\lambda t}  - (\lambda t)^2 \\ 
  &= \lambda t \Bigg( \sum_{i=0}^\infty \Big( \frac{(\lambda t)^{n-1}}{(n-2)!}e^{-\lambda t} + \frac{(\lambda t)^{n-1}}{(n-1)!}e^{-\lambda t} \Big) \Bigg)  - (\lambda t)^2 \\ 
  &= \lambda t \Big( \lambda t \sum_{i=0}^\infty \frac{(\lambda t)^{n-2}}{(n-2)!}e^{-\lambda t} + \sum_{i=0}^\infty \frac{(\lambda t)^{n-1}}{(n-1)!}e^{-\lambda t} \Big)  - (\lambda t)^2 \\ 
  &= \lambda t (\lambda t + 1)  - (\lambda t)^2 \\ 
  &= \lambda t \rightarrow \frac{\lambda t}{100} \textrm{for t in units of cM}
\end{align}

### 2.
*Given that a single crossover occurs along a chromosome of length $t$, provide
a proof based on the basic properties of a Poisson Process that the location of
the crossover is uniformly distributed on $(0, t]$.*

For $0 < s < t$,

\begin{align}
P(S < s | N_t = 1) &= \frac{P(S < s, N_t=1)}{P(N_t=1)} \\
  &= \frac{P(N_s=1, N_t-N_s=0)}{P(N_t=1)} \\
  &= \frac{\lambda s e^{-\lambda s} e^{-\lambda(t-s)}}{\lambda t e^{-\lambda t}} \\
  &= \frac{s}{t} \\
\implies p(S = s | N_t = 1) &= \frac{d}{ds}\frac{s}{t} \\
  &= \frac{1}{t}
\end{align}

The distribution for $P(S < s | N_t = 1)$ is the CDF of a uniform distribution
over $(0,t]$, so $S \sim \textrm{Unif}(0,t)$.


### 3.
*For a chromosome of length 3 Morgans (e.g. chromosome one in humans):*

* *What is the probability that there are no crossovers? Define the random
  variable $C$ as the number of crossovers.*

\begin{align}
P(C = 0) &= P(N(t)=0) = e^{-3}
\end{align}

* *What is the expected number of crossovers, $\mathbb{E}[C]$?*

\begin{align}
E[C] &= E[N(t)] = 3
\end{align}

* *What is the probability of $\mathbb{E}[C]$ or more crossovers?*

\begin{align}
P(C > E[C]) &= P(C > 3) \\
  &= \sum_{n=3}^\infty \frac{3^n e^{-3}}{n!} \\
  &= 1 - \sum_{n=0}^2 \frac{3^n e^{-3}}{n!} \\
  &= 1 - e^{-3} - \frac{3^1}{1!}e^{-3} - \frac{3^2}{2!} e^{-3} \\
  &= 1 - e^{-3} - 3e^{-3} - \frac{9}{2} e^{-3} \\
  &= 1 - 8.5 e^{-3} \\
  &= 0.5768099
\end{align}

### 4.
*Consider a location with a new mutation on a chromosome. As that chromosome is
passed down through $k$ rounds of meiosis, what is the distribution to the
distance to the first crossover to the right of the mutation?*

Assuming the location $s$ of the new mutation is uniformly distributed over the
chromosome of length $L$, the distribution of crossovers on either side of the
mutation will both be Poisson distributed thinning processes with parameters
$s$ and $L-s$, and thereby have a joint distribution $\sim Poi(L)$.

The distribution for $k$ crossovers is the sum of $k$ Poisson processes, which
produces a new Poisson process $\sim Poi(kL)$.

The distribution for the waiting 'time' between the mutation and the first
crossover event to its right is as follows:

\begin{align}
P(T > t) &= P(N(t) = 0) \\
  &= \frac{k(L-s)^0}{0!}e^{-k(L-s)t} \\
  &= e^{-k(L-s)t} \\
\implies P(T < t) &= 1 - e^{-k(L-s)t} \\
\implies p(T=t ) &= \frac{d}{dt} \big( 1-e^{-k(L-s)t} \big) \\
  &= k(L-s)e^{-k(L-s)t}
\end{align}

This is exponentially distributed with rate parameter $k(L-s)$.

The expected value for this distribution is $\frac{1}{k(L-s)}$, the expectation
for an exponential. This number, the expected crossover site per unit length,
multiplied by the distance the right of the mutation, $L-s$, yields:

\begin{align}
E[distance] &= \frac{1}{k}
\end{align}

Also, consider the following simulation of crossovers. Plotted are simulated
expected values for the distances to nearest crossover sites (right or left)
for various values of $k$.

```{r}
runMeiosis <- function(N, k, L=1, x=0.5) {
  r <- matrix(0, N, k) # N replicates for 1:k values for k
  l <- matrix(0, N, k)
  for (i in 1:N) {
    for (j in 1:k) {
      d <- runif(L*j, 0, L) - x
      r[i,j] <- min(d[d>0])
      l[i,j] <- max(d[d<0])

      # Add NAs so that mean() will ignore columns where the number of
      # crossovers on the left or right side is 0.
      if (is.infinite(r[i,j])) {
        r[i,j] <- NA
      }
      if (is.infinite(l[i,j])) {
        l[i,j] <- NA
      }
    }
  }
  return(list(r=r,l=l))
}
```

```{r, warning=FALSE}
N <- 100
crossoverRate <- 100
L <- 10
x0 <- 4
res <- runMeiosis(N, crossoverRate, L, x0)
r <- res$r
l <- res$l

k <- seq(1,crossoverRate)

plot(k, apply(r, 2, mean))
lines(k, 1/k)
```


### 5.
*Consider the distance between the first crossover to the left of the mutation
and the first crossover to the right of the mutation. What is the expected
length of this distance as function of $k$?*

First, evaluate the expected distance for the left. This follows a similar
argument to that of the right, but with the exponential rate parameter $ks$
instead of $k(L-s)$:

\begin{align}
P(T > t) &= P(N(t) = 0) \\
  &= \frac{ks^0}{0!}e^{-kst} \\
  &= e^{-kst} \\
\implies P(T < t) &= 1 - e^{-kst} \\
\implies p(T=t ) &= \frac{d}{dt} \big(1 - e^{-kst} \big) \\
  &= kse^{-kst}
\end{align}

The distance is exponentially distributed with rate parameter $ks$.

The expected value for this distribution is $\frac{1}{ks}$, the expectation for
an exponential. This number, multiplied by the distance to the left of the
mutation, $s$, yields:

\begin{align}
E[distance] &= \frac{1}{k}
\end{align}

Consider the simulation for left-side mutations:

```{r}
plot(k, apply(-l, 2, mean))
lines(k, 1/k)
```

Given then that the two intervals $(0,s]$ and $(s,L]$ are independent, the
expectation of the distance between the two is equal sum of the expectations of
the distances of the two.

\begin{align}
E[right + left] &= E[right] + E[left] \\
  &= \frac{1}{k} + \frac{1}{k} \\
  &= \frac{2}{k}
\end{align}

See the plot for the mean distances for various values of $k$ below.

```{r}
plot(k, apply(r-l, 2, mean))
lines(k, 2/k)
```

### 6.
*If an odd number of crossovers occur between two locations separated by
distance $d$ on a chromosome then a recombination event occurs. Calculate the
probability of at least one recombination event occurring as a function of
$d$.*

\begin{align}
P(\textrm{at least 1 recombination}) &= 1 - P(\textrm{no recombinations}) \\
  &= 1 - P(\textrm{even no. of crossovers})
\end{align}

\begin{align}
P(\textrm{even}) &= e^{-d} \Big(\frac{d^0}{0!} + \frac{d^2}{2!} + \frac{d^4}{4!} + \frac{d^6}{6!} + \dots \Big) \\
  &= e^{-d} \sum_{m=0}^{\infty} \frac{d^{2m}}{2m!} \\
  &= e^{-d} \Big( \frac{e^d + e^{-d}}{2} \Big) \\
  &= \frac{1 + e^{-2d}}{2} \\
\implies P(odd) &= 1 - \frac{1 + e^{-2d}}{2} \\
  &= \frac{1 - e^{-2d}}{2}
\end{align}


## Cancer cell population model

*Background: Consider modeling cancer cells entering the blood/circulatory
systems as a non-homogeneous Poisson Process. In particular, suppose there is a
source cancer cell population whose size is growing deterministically. More
specifically, assume $x(t)$, the size of the source population at time $t$, is
growing exponentially at rate $r$ per hour (i.e. $x(t) = x_0 e^{rt}$ ) where
$x_{0}$ is the starting size of the population. An expected fraction $m$ (for $0
\leq m \leq 1$) of that cancer population is entering the blood/circulatory
system, such that cancer cells enter the blood as a Poisson Process with an
instantaneous rate of $mx(t)$ per hour.*

*Answer the following questions:*

### 1. 
*Provide an expression for the probability distribution on the number of cancer
cells in the blood after $t$ hours.*

Given $m(t)$:

\begin{align}
m(t) &= \int_{0}^t mx_{0}e^{rs} ds \\
  &= \frac{mx_{0}}{r}\big(e^{rt} -1 \big) \\
\end{align}

The probability of $n$ cancer cells at time $t$ is given by the nonhomogeneous
Poisson distribution:

\begin{align}
P(N(t)=n) &= \frac{{m(t)}^n}{n!} e^{m(t)} \\
  &= \frac{{\Big(\frac{mx_{0}}{r}\big(e^{rt} -1 \big)\Big)}^n}{n!} e^{\frac{mx_{0}}{r}\big(e^{rt} -1 \big)}
\end{align}


### 2.
*If $x_0 = 10$, $r = 0.01$, and $m = 0.0001$, at what time is there a 99\%
probability that the blood will have received 1000 cancer cells?*

To solve this unwieldy equation for $t$ when $n=1000$ and $P(N(t)=1000)=.99$,
solve for the roots of the expression $P(N(t)=1000)-.99 = 0$.

Due to numerical issues, the solution is approximated first by using `qpois` to
solve for the rate parameter that corresponds to $P(N(t)=1000)=.99$.

```{r}
mt <- function(t) {
  x0 <- 10
  r <- .01
  m <- .0001
  return((m*x0/r)*(exp(r*t)-1))
}

findZero <- function(t) {
  return(ppois(1000, t, lower.tail=F)-.99)
}

logFactorial <- function(x) {
  return(sum(log(seq(1,x))))
}
```

```{r}
x0 <- 10
r <- .01
m <- .0001

root <- uniroot(findZero, c(1,2000))$root

log((root*r/x0/m)+1)/r
```

The value of $t \approx 928.3748$ hours.


### 3.
*Suppose that the probability a cancer cell in the blood is still living at time
$t$ is determined by an exponential function of its arrival time to the blood
and a death rate of $\mu$ per hour and, i.e.*

\begin{align}
P{\textrm{living}}(t) = e^{-\mu t}
\end{align}

*Provide a general form for the number of living cancer cells in the blood at
time $t$.*

\begin{align}
m(t) &= \int_{0}^t mx_{0}e^{rs} e^{-\mu (t-s)} ds \\
  &= mx_0 e^{-\mu t} \int_{0}^t e^{rs} e^{\mu(s)} ds \\
  &= mx_0 e^{-\mu t} \int_{0}^t e^{(r+\mu)s} ds \\
  &= \frac{mx_0}{r+\mu} e^{-\mu t} \big( e^{(r+\mu)t} - 1 \big) \\
  &= \frac{mx_0}{r+\mu} \big( e^{rt} - e^{-\mu t} \big)
\end{align}

The probability distribution follows the same Poisson type of distribution, but
it is even more unwieldy:

\begin{align}
P(N(t)=n) &= \frac{{m(t)}^n}{n!} e^{m(t)} \\
  &= \frac{{\Big( \frac{mx_0}{r+\mu} \big( e^{rt} - e^{-\mu t} \big) \Big)}^n}{n!} e^{ \frac{mx_0}{r+\mu} \big( e^{rt} - e^{-\mu t} \big) }
\end{align}


### 4.
Repeat part 2 but address when the blood will have 1000 living cancer cells.
Assume $\mu = 0.01$.

```{r}
mt2 <- function(t) {
  x0 <- 10
  r <- .01
  m <- .0001
  mu <- .01
  return(m*x0/(r+mu) * (exp((r)*t) - exp(-mu*t) ))
}

mt2zero <- function(t) {
  x0 <- 10
  r <- .01
  m <- .0001
  mu <- .01
  
  # 1076.07 is the value of lambda that is > 1000 with probability .99. Find
  # the root for the exponential.
  return(m*x0/(r+mu) * (exp((r)*t) - exp(-mu*t) ) - 1076.07)
}
```

```{r}
x0 <- 10
r <- .01
m <- .0001
mu <- .01

# uniroot(mt2zero, c(1,2000))$root
100*asinh(root*(r+mu)/m/x0/2)
```

$t \approx 997.6803$ hours in this scenario.


### 5.
*An assumption of a Poisson Process is that arrivals are rare enough that at
most one arrival is expected in a given $\Delta t$ unit of time as $\Delta t$
goes to zero. Keeping this in mind --- when might we expect the Poisson Process
to fail as an approximation for modeling cancer cell counts in the blood? (A
conceptual/verbal answer is okay, extra credit for a more precise answer).*

Several assumptions must be met for the Poisson approximation to hold.

(i) $N(0) = 0$

(ii) $N(t), t \geq 0$ has stationary and independent increments

(iii) $P[ N(t + h) - N(t) \geq 2 ] = o(h)$

(iv) $P[ N(t + h) - N(t) = 1 ] = \lambda h + o(h)$

These assumptions can be violated if, for example:

 * Cancer cell counts are contingent on prior population history, possible due
   to immune system response. Violates (ii).
 * The fraction of cells entering the blood is high relative to the growing
   population. Violates (iii, iv).
 * Cells are not released into the bloodstream as a random Bernoulli trial with
   low probability of emission and large sample size. Violates (iii, iv).
 * There are already cancer cells in the bloodstream. Violates (i).
 * The initial tumor population size in the tissue is small.

In the above problems, the emission rate $m$ for a given population size $x(t)$
is low, so the Poisson approximation is valid. The true distribution would
treat the population x(t) as finite and model every emission into the
bloodstream as also a loss in the tumor population. To test the goodness of fit
for the Poisson model, one would use a likelihood ratio test ($\chi^2$)
comparing true values of the population sizes or, if not available, data
sampled from populations.


# Project Ideas

*Write a paragraph describing your current idea for the final project.*

*If you are planning to do a research project:*

 * *Define the biological problem or question that you are addressing.*
 * *Outline (broadly) the statistical models and/or algorithms that you will
   use to address your biological problem or question.*

*If you are planning to provide a teaching resource:*

 * *Define the subject area(s) that you will address.*
 * *Outline (broadly) materials that you will develop.*


I have a couple ideas. They are still inchoate at best.


## Idea 1 (research?)

I happened upon *The Population Consequences of Life History Phenomena*, and
it's giving me some ideas.

The broad question is what are the dynamics of changes populations
(birth/death) can be modelled as continuous-time Markov chains. My thinking is
to:

* Build matrix representations for continuous time Markov chains for
  populations in an ecosystem.

* Many of the equations use an odd notation, and I think they would be very
  amenable to conversion to forms we learned in lecture and subsequent
  analysis (eigendecomposition, exploiting matrix structure, etc).

I wouldn't really call this research, as I suspect many, many theoretical
ecologist have already built and published more sophisticated models than I
know of and that run more efficiently than what I would make. My motivation
doing this is to become better at reading, implementing, and ultimately
understanding matrix notations for complex systems. I don't have a particular
research question I want to investigate or any data, for that matter.


## Idea 2 (teaching resource)

Add some vignettes to the website on how to deal with numerical errors. Not
really the purview of the course, but on many problems, we had to adjust our
computations to deal with under/overflow or simply abandon self-coded methods
and look for other (likely system/package functions) ones that somehow
successfully compute what we need.

Topics to cover:

 * Overview of why numerical errors accumulate
 * Strategies for accounting for them: by hand (renormalizing,
   log-transformation)
 * Using system functions instead (I don't grok R, so I will have to look
   things up and hope that I can understand...)
 * Assessing errors in computations

Numerical stability and other low-level topics are well outside my wheelhouse,
so I don't know now how well I can pull this off. All I know is how annoying it
is to run into numerical errors while coding up solutions and trying to figure
out how to control them.
