---
title: Project
author: Ansel George
output: pdf_document
---

```{r}
library(expm)
library(ggplot2)
library(dplyr)
library(tidyr)

set.seed(10)
```


# Define functions

```{r}
# Use a beta function to generate survival probabilities for all states. Use
# pbeta for discretization instead of dbeta to get the area between intervals.
makeDeathRates <- function(t, a, w, d, d_a=1, d_b=1) {
  lambda <- max(t)+1
  p <- pbeta((t+1)/lambda, d_a, d_b) - pbeta(t/lambda, d_a, d_b)
  p <- 1-p
  p <- p/mean(p)
  p <- p*(1-d)
  return(p)
}

# Generate the survival matrix using a survival function and the dimensions
# specified by the demographic parameters of the population.
#   fxn - fitness function
#   tstep - time step between stages in the state chain
#   lambda - longevity
#   a - age of start of fertility
#   w - age of end of fertility
#   d - death probability for moving from stage to stage
#   d_a - alpha parameter for the beta function
#   d_b - beta parameter for the beta function
makeSurvivalMatrix <- function(fxn, tstep, lambda, a, w, d=.1, d_a=1, d_b=1) {
  return(rbind(cbind(rep(0, lambda-1),
                     diag(lambda-1)),
               rep(0, lambda)) * fxn(seq(0, (lambda-1)*tstep), a, w, d, d_a, d_b))
}

# Uniform litter size
flatBirthRates <- function(t, a, w, m, v) {
  b <- rep(0, length(t))
  b[a:w] <- 1
  return(b)
}

# Normal-shaped (due to discretization) litter size with age. It is scaled so
# that the total fecundity for individuals (assuming survival) is the same as
# that of the uniform litter size function.
normalBirthRates <- function(age, a, w, m, v) {
  p <- dnorm(age, m, sqrt(v))
  return((w-a+1)*p/sum(p))
}

# Use the birth function and population parameters to make the birth matrix.
#   fxn - litter size distribution function
#   tstep - time step size
#   lambda - longevity
#   a - age of fertility (for uniform)
#   w - age of end of fertility (for uniform)
#   m - mean of normal litter size distribution (for normal)
#   v - variance of litter size (for normal)
#   b - litter size
makeBirthMatrix <- function(fxn, tstep, lambda, a=0, w=0, m=1, v=0, b=1) {
  B <- matrix(0, nrow=lambda, ncol=lambda)
  B[1,] <- b*fxn(seq(0, (lambda-1)*tstep, tstep), a, w, m, v)
  return(t(B))
}

# Modify makeBirthMatrix to penalize large litter sizes.
#   fxn - litter size distribution function
#   tstep - time step size
#   lambda - longevity
#   a - age of fertility (for uniform)
#   w - age of end of fertility (for uniform)
#   m - mean of normal litter size distribution (for normal)
#   v - variance of litter size (for normal)
#   b - litter size
#   s - strength of constraint against litter size
makeConstrainedBirthMatrix <- function(fxn, tstep, lambda, a=0, w=0, m=1, v=0, b=1, s=1) {
  B <- matrix(0, nrow=lambda, ncol=lambda)
  # B_survive <- diag(pmin(1, (seq(0,lambda-1)/b)^s))
  B_survive <- diag((rep(1, lambda)/b)^s)
  B[1,] <- b*fxn(seq(0, (lambda-1)*tstep, tstep), a, w, m, v) %*% B_survive
  return(t(B))
}

# Unused, eigenvalues are a much better solution...
simulateGenerations <- function(x, P, B, n) {
  x_new <- x %*% ((P+B) %^% n)
  x_new <- x_new / sum(x_new)
  return(x_new)
}
```


# Global variables

```{r}
tstep <- 1   # timestep
a <- 20      # fertile age
w <- 30      # end fertile age
b <- 5       # litter size
d <- .1      # death change per timestep
m <- 20      # mean normal birth function
v <- 5       # var normal birth function
s <- 1       # constraint
lambda <- 50 # longevity
```


# Simple simulations

## Uniform birth/death rates

```{r}
# code for faffing about...
P <- makeSurvivalMatrix(makeDeathRates, tstep, lambda, a, w, d, 1, 1)
# B <- makeConstrainedBirthMatrix(flatBirthRates, tstep, lambda, a, w, m, v, b)
B <- makeBirthMatrix(flatBirthRates, tstep, lambda, a, w, m, v, b)

res<-eigen(t(P+B))
log(res$values[1])
x_pi <- res$vectors[,1] / sum(res$vectors[,1])

age <- seq(1, lambda, tstep)
```

### Lambda

Plot different values of lambda (longevity) holding other things constant.

```{r}
lambdas <- seq(w, 5*w, w)
df_lambda <- matrix(0, nrow=length(lambdas), ncol=5*w)
df_lambda_r <- matrix(0, nrow=length(lambdas), ncol=1)

for (i in 1:length(lambdas)) {
  P <- makeSurvivalMatrix(makeDeathRates, tstep, lambdas[i], a, w, d, 1, 1) 
  B <- makeConstrainedBirthMatrix(flatBirthRates, tstep, lambdas[i], a, w, m, v, b, s)
  res<-eigen(t(P+B))
  df_lambda_r[i] <- log(Re(res$values[1]))
  x_pi <- res$vectors[,1] / sum(res$vectors[,1])
  # df_lambda[i,] <- c(Re(x_pi), rep(0, 5*w-lambdas[i]))
  df_lambda[i,] <- c(Re(x_pi), rep(NA, 5*w-lambdas[i]))
}

header <- paste('lambda', lambdas, sep=' ')
rownames(df_lambda) <- header
df_lambda <- tbl_df(t(df_lambda))
df_lambda <- df_lambda %>% mutate("age"=seq(1, 5*w))
f1 <- df_lambda %>% gather(lambda, freq, -age) %>% ggplot() + aes(x=age,y=freq,color=lambda) + geom_line()
ggsave(file="freq_lambda-1-1.svg", plot=f1)

# Plot growth rate given several values of lambda
f2 <- qplot(lambdas, df_lambda_r, geom='line')
ggsave(file="r_lambda-1-1.svg", plot=f2)
```


### Alpha

Plot different values of alpha (first reproduction age) holding other things
constant.

```{r}
alphastep <- 1
alphas <- seq(1, (w-1), alphastep)
df_alpha <- matrix(0, nrow=length(alphas), ncol=lambda)
df_alpha_r <- matrix(0, nrow=length(alphas), ncol=1)
for (i in 1:length(alphas)) {
  P <- makeSurvivalMatrix(makeDeathRates, tstep, lambda, alphas[i], w, d, 1, 1)
  B <- makeConstrainedBirthMatrix(flatBirthRates, tstep, lambda, alphas[i], w, m, v, b, s)
  res<-eigen(t(P+B))
  df_alpha_r[i] <- log(Re(res$values[1]))
  x_pi <- res$vectors[,1] / sum(res$vectors[,1])
  df_alpha[i,] <- Re(x_pi)
}
header <- paste('alpha=0', alphas[1:ceiling(9/alphastep)], sep='')
header <- c(header, paste('alpha=', alphas[-(1:ceiling(9/alphastep))], sep=''))

rownames(df_alpha) <- header
df_alpha <- tbl_df(t(df_alpha))

df_alpha <- df_alpha %>% mutate("age"=seq(1,lambda))
f1 <- df_alpha %>%
  gather(alpha, freq, -age) %>%
  ggplot() + aes(x=age,y=freq,color=alpha) + geom_line()
ggsave(file="freq_alpha-1-1.svg", plot=f1, width=6, height=4)

# Plot extra curves for alpha given a different constraint s
df_alpha_r2 <- matrix(0, nrow=length(alphas), ncol=1)
for (i in 1:length(alphas)) {
  P <- makeSurvivalMatrix(makeDeathRates, tstep, lambda, alphas[i], w, d, 1, 1)
  B <- makeConstrainedBirthMatrix(flatBirthRates, tstep, lambda, alphas[i], w, m, v, b, 2)
  res<-eigen(t(P+B))
  df_alpha_r2[i] <- log(Re(res$values[1]))
}

df_alpha_r3 <- matrix(0, nrow=length(alphas), ncol=1)
for (i in 1:length(alphas)) {
  P <- makeSurvivalMatrix(makeDeathRates, tstep, lambda, alphas[i], w, d, 1, 1)
  B <- makeConstrainedBirthMatrix(flatBirthRates, tstep, lambda, alphas[i], w, m, v, b, .5)
  res<-eigen(t(P+B))
  df_alpha_r3[i] <- log(Re(res$values[1]))
}

tmp<-data.frame(df_alpha_r, df_alpha_r2, df_alpha_r3)
colnames(tmp) <- c('s=1', 's=2', 's=0.5')
df_tmp <- tbl_df(tmp) %>% mutate('alpha'=alphas)
f <- df_tmp %>% gather(s, r, -alpha) %>% ggplot() + aes(x=alpha, y=r, color=s) + geom_line()
ggsave(file="r_alpha-1-1_extra.svg", plot=f, width=6, height=4)
```

### Omega

Plot different 'menopause' distribution (omega) holding other things constant.

```{r}
omegastep <- 2
omegas <- seq(a+1, lambda, omegastep)
df_omega <- matrix(0, nrow=length(omegas), ncol=lambda)
df_omega_r <- matrix(0, nrow=length(omegas), ncol=1)
for (i in 1:length(omegas)) {
  P <- makeSurvivalMatrix(makeDeathRates, tstep, lambda, a, w, d, 1, 1)
  B <- makeConstrainedBirthMatrix(flatBirthRates, tstep, lambda, a, omegas[i], m, v, b, s)
  res<-eigen(t(P+B))
  df_omega_r[i] <- log(Re(res$values[1]))
  x_pi <- res$vectors[,1] / sum(res$vectors[,1])
  df_omega[i,] <- Re(x_pi)
}
df_omega_r2 <- matrix(0, nrow=length(omegas), ncol=1)
for (i in 1:length(omegas)) {
  P <- makeSurvivalMatrix(makeDeathRates, tstep, lambda, a, w, d, 1, 1)
  B <- makeConstrainedBirthMatrix(flatBirthRates, tstep, lambda, a, omegas[i], m, v, b, 2)
  res<-eigen(t(P+B))
  df_omega_r2[i] <- log(Re(res$values[1]))
}
df_omega_r3 <- matrix(0, nrow=length(omegas), ncol=1)
for (i in 1:length(omegas)) {
  P <- makeSurvivalMatrix(makeDeathRates, tstep, lambda, a, w, d, 1, 1)
  B <- makeConstrainedBirthMatrix(flatBirthRates, tstep, lambda, a, omegas[i], m, v, b, 0.5)
  res<-eigen(t(P+B))
  df_omega_r3[i] <- log(Re(res$values[1]))
}
header <- paste('omega=', omegas, sep='')
rownames(df_omega) <- header
df_omega <- tbl_df(t(df_omega))

df_omega <- df_omega %>% mutate("age"=seq(1,lambda))
f1 <- df_omega %>% gather(omega, freq, -age) %>% ggplot() + aes(x=age,y=freq,color=omega) + geom_line()
ggsave(file="freq_omega-1-1.svg", plot=f1, width=6, height=4)

tmp<-data.frame(df_omega_r, df_omega_r2, df_omega_r3)
colnames(tmp) <- c('s=1', 's=2', 's=0.5')
df_tmp <- tbl_df(tmp) %>% mutate('omega'=omegas)
f <- df_tmp %>% gather(s, r, -omega) %>% ggplot() + aes(x=omega, y=r, color=s) + geom_line()
ggsave(file="r_omega-1-1_extra.svg", plot=f, width=6, height=4)
```

### b (litter size)

Plot different 'litter size' distribution (b) holding other things constant.

```{r}
bstep <- 1
bs <- seq(1, 10, bstep)
df_b <- matrix(0, nrow=length(bs), ncol=lambda)
df_b_r <- matrix(0, nrow=length(bs), ncol=1)
for (i in 1:length(bs)) {
  P <- makeSurvivalMatrix(makeDeathRates, tstep, lambda, a, w, d, 1, 1)
  B <- makeConstrainedBirthMatrix(flatBirthRates, tstep, lambda, a, w, m, v, bs[i], 1)
  res<-eigen(t(P+B))
  df_b_r[i] <- log(Re(res$values[1]))
  x_pi <- res$vectors[,1] / sum(res$vectors[,1])
  df_b[i,] <- Re(x_pi)
}
df_b_r2 <- matrix(0, nrow=length(bs), ncol=1)
for (i in 1:length(bs)) {
  P <- makeSurvivalMatrix(makeDeathRates, tstep, lambda, a, w, d, 1, 1)
  B <- makeConstrainedBirthMatrix(flatBirthRates, tstep, lambda, a, w, m, v, bs[i], 2)
  res<-eigen(t(P+B))
  df_b_r2[i] <- log(Re(res$values[1]))
}
df_b_r3 <- matrix(0, nrow=length(bs), ncol=1)
for (i in 1:length(bs)) {
  P <- makeSurvivalMatrix(makeDeathRates, tstep, lambda, a, w, d, 1, 1)
  B <- makeConstrainedBirthMatrix(flatBirthRates, tstep, lambda, a, w, m, v, bs[i], .5)
  res<-eigen(t(P+B))
  df_b_r3[i] <- log(Re(res$values[1]))
}
header <- paste('b=', bs, sep='')
rownames(df_b) <- header
df_b <- tbl_df(t(df_b))

df_b <- df_b %>% mutate("age"=seq(1,lambda))
f1 <- df_b %>% gather(b, freq, -age) %>% ggplot() + aes(x=age,y=freq,color=b) + geom_line()
ggsave(file="freq_b-1-1.svg", plot=f1, width=6, height=4)

tmp<-data.frame(df_b_r, df_b_r2, df_b_r3)
colnames(tmp) <- c('s=1', 's=2', 's=0.5')
df_tmp <- tbl_df(tmp) %>% mutate('b'=bs)
f <- df_tmp %>% gather(s, r, -b) %>% ggplot() + aes(x=b, y=r, color=s) + geom_line()
ggsave(file="r_b-1-1_extra.svg", plot=f, width=6, height=4)
```

### s (constraint)

Plot different 'strength of constraint' (s) holding other things constant.

```{r}
sstep <- .1
ss <- seq(0, 5, sstep)
df_s <- matrix(0, nrow=length(ss), ncol=lambda)
df_s_r <- matrix(0, nrow=length(ss), ncol=1)
for (i in 1:length(ss)) {
  P <- makeSurvivalMatrix(makeDeathRates, tstep, lambda, a, w, d, 1, 1)
  B <- makeConstrainedBirthMatrix(flatBirthRates, tstep, lambda, a, w, m, v, b, ss[i])
  res<-eigen(t(P+B))
  df_s_r[i] <- log(Re(res$values[1]))
  x_pi <- res$vectors[,1] / sum(res$vectors[,1])
  df_s[i,] <- Re(x_pi)
}
df_s_r2 <- matrix(0, nrow=length(ss), ncol=1)
for (i in 1:length(ss)) {
  P <- makeSurvivalMatrix(makeDeathRates, tstep, lambda, a, w, d, 1, 1)
  B <- makeConstrainedBirthMatrix(flatBirthRates, tstep, lambda, a, w, m, v, 1, ss[i])
  res<-eigen(t(P+B))
  df_s_r2[i] <- log(Re(res$values[1]))
}
df_s_r3 <- matrix(0, nrow=length(ss), ncol=1)
for (i in 1:length(ss)) {
  P <- makeSurvivalMatrix(makeDeathRates, tstep, lambda, a, w, d, 1, 1)
  B <- makeConstrainedBirthMatrix(flatBirthRates, tstep, lambda, a, w, m, v, 10, ss[i])
  res<-eigen(t(P+B))
  df_s_r3[i] <- log(Re(res$values[1]))
}
header <- paste('s=', ss, sep='')
rownames(df_s) <- header
df_s <- tbl_df(t(df_s))

df_s <- df_s %>% mutate("age"=seq(1,lambda))
f1 <- df_s %>% gather(s, freq, -age) %>% ggplot() + aes(x=age,y=freq,color=s) + geom_line()
ggsave(file="freq_s-1-1.svg", plot=f1, width=6, height=4)

tmp<-data.frame(df_s_r, df_s_r2, df_s_r3)
colnames(tmp) <- c('b=5', 'b=1', 'b=10')
df_tmp <- tbl_df(tmp) %>% mutate('s'=ss)
f <- df_tmp %>% gather(b, r, -s) %>% ggplot() + aes(x=s, y=r, color=b) + geom_line()
ggsave(file="r_s-1-1_extra.svg", plot=f, width=6, height=4)
```

### d (Death)

Plot different death distribution holding other things constant.

```{r}
deaths <- seq(0, .45, .025)
df_death <- matrix(0, nrow=length(deaths), ncol=lambda)
df_death_r <- matrix(0, nrow=length(deaths), ncol=1)
for (i in 1:length(deaths)) {
  P <- makeSurvivalMatrix(makeDeathRates, tstep, lambda, a, w, deaths[i], 1, 1)
  B <- makeConstrainedBirthMatrix(flatBirthRates, tstep, lambda, a, w, m, v, b)
  res<-eigen(t(P+B))
  df_death_r[i] <- log(Re(res$values[1]))
  x_pi <- res$vectors[,1] / sum(res$vectors[,1])
  df_death[i,] <- Re(x_pi)
}
header <- paste('death', deaths, sep=' ')
rownames(df_death) <- header
rownames(df_death) <- header
df_death <- tbl_df(t(df_death))

df_death <- df_death %>% mutate("age"=seq(1,lambda))
df_death %>%
  gather(death, freq, -age) %>%
  ggplot() + aes(x=age,y=freq,color=death) + geom_point()

qplot(deaths, df_death_r, geom='line')
```

## Normal birth, flat death

### Lambda

Plot different values of lambda (longevity) holding other things constant.

```{r}
lambdas <- seq(w, 5*w, w)
df_lambda <- matrix(0, nrow=length(lambdas), ncol=5*w)
df_lambda_r <- matrix(0, nrow=length(lambdas), ncol=1)

for (i in 1:length(lambdas)) {
  P <- makeSurvivalMatrix(makeDeathRates, tstep, lambdas[i], a, w, d, 1, 1) 
  B <- makeConstrainedBirthMatrix(normalBirthRates, tstep, lambdas[i], a, w, m, v, b, s)
  res<-eigen(t(P+B))
  df_lambda_r[i] <- log(Re(res$values[1]))
  x_pi <- res$vectors[,1] / sum(res$vectors[,1])
  # df_lambda[i,] <- c(Re(x_pi), rep(0, 5*w-lambdas[i]))
  df_lambda[i,] <- c(Re(x_pi), rep(NA, 5*w-lambdas[i]))
}

header <- paste('lambda', lambdas, sep=' ')
rownames(df_lambda) <- header
df_lambda <- tbl_df(t(df_lambda))
df_lambda <- df_lambda %>% mutate("age"=seq(1, 5*w))
f1 <- df_lambda %>% gather(lambda, freq, -age) %>% ggplot() + aes(x=age,y=freq,color=lambda) + geom_line()
ggsave(file="freq_norm_lambda-1-1.svg", plot=f1)

# Plot growth rate given several values of lambda
f2 <- qplot(lambdas, df_lambda_r, geom='line')
ggsave(file="r_norm_lambda-1-1.svg", plot=f2)
```

### m (mean)

Plot different values of mean reproduction age (m) holding other things
constant.

```{r}
meanstep <- 1
means <- seq(ceiling(m/2), 2*m, meanstep)
df_mean <- matrix(0, nrow=length(means), ncol=lambda)
df_mean_r <- matrix(0, nrow=length(means), ncol=1)
for (i in 1:length(means)) {
  P <- makeSurvivalMatrix(makeDeathRates, tstep, lambda, a, w, d, 1, 1) 
  B <- makeConstrainedBirthMatrix(normalBirthRates, tstep, lambda, a, w, means[i], v, b, s)
  res<-eigen(t(P+B))
  df_mean_r[i] <- log(Re(res$values[1]))
  x_pi <- res$vectors[,1] / sum(res$vectors[,1])
  df_mean[i,] <- Re(x_pi)
}
header <- paste('m=', means, sep='')

rownames(df_mean) <- header
colnames(df_mean_r) <- 'r'
df_mean <- tbl_df(t(df_mean))

df_mean <- df_mean %>% mutate("age"=seq(1,lambda))
df_mean %>%
  gather(m, freq, -age) %>%
  ggplot() + aes(x=age,y=freq,color=m) + geom_line()

df_mean_r <- tbl_df(df_mean_r) %>% mutate(m=means)
df_mean_r %>%
  ggplot() + aes(x=m, y=r) + geom_line() + geom_hline(aes(yintercept=0), color='red')
```


### v (variance)

Plot variance of different distributions other things constant.

```{r}
vstep <- 1
vs <- seq(2, 10)
t_v <- matrix(0, nrow=length(vs), ncol=lambda)
t_v_r <- matrix(0, nrow=length(vs), ncol=1)
for (i in 1:length(vs)) {
  P <- makeSurvivalMatrix(makeDeathRates, tstep, lambda, a, w, d, 1, 1) 
  B <- makeConstrainedBirthMatrix(normalBirthRates, tstep, lambda, a, w, m, vs[i], b, s)
  res <- eigen(t(P+B))
  t_v_r[i] <- log(Re(res$values[1]))
  x_pi <- res$vectors[,1] / sum(res$vectors[,1])
  t_v[i,] <- Re(x_pi)
}
header <- paste('v=0', vs[1:8], sep='')
header <- c(header, paste('v=', vs[-(1:8)], sep=''))

rownames(t_v) <- header
colnames(t_v_r) <- 'r'
df_v <- tbl_df(t(t_v))

df_v <- df_v %>% mutate("age"=seq(1,lambda))
df_v %>%
  gather(v, freq, -age) %>%
  ggplot() + aes(x=age,y=freq,color=v) + geom_line()

df_v_r <- tbl_df(t_v_r) %>%
  mutate(v=vs)
df_v_r %>%
  ggplot() + aes(x=v, y=r) + geom_line() + geom_hline(aes(yintercept=0), col='red')
```

### d (death)

Plot different death distribution holding other things constant.

```{r}
deaths <- seq(0, .45, .025)
df_death <- matrix(0, nrow=length(deaths), ncol=lambda)
df_death_r <- matrix(0, nrow=length(deaths), ncol=1)
for (i in 1:length(deaths)) {
  P <- makeSurvivalMatrix(makeDeathRates, tstep, lambda, a, w, deaths[i], 1, 1)
  B <- makeConstrainedBirthMatrix(normalBirthRates, tstep, lambda, a, w, m, v, b)
  res<-eigen(t(P+B))
  df_death_r[i] <- log(Re(res$values[1]))
  x_pi <- res$vectors[,1] / sum(res$vectors[,1])
  df_death[i,] <- Re(x_pi)
}

header <- paste('death', deaths, sep=' ')
rownames(df_death) <- header
rownames(df_death) <- header
df_death <- tbl_df(t(df_death))

df_death <- df_death %>% mutate("age"=seq(1,lambda))
df_death %>% 
  gather(death, freq, -age) %>% 
  ggplot() + aes(x=age,y=freq,color=death) + geom_point()

qplot(deaths, df_death_r, geom='line')
```


## High young mortality, flat birth rate 

### Lambda

Plot different values of lambda (longevity) holding other things constant.

```{r}
lambdas <- seq(w, 5*w, w)
df_lambda <- matrix(0, nrow=length(lambdas), ncol=5*w)
df_lambda_r <- matrix(0, nrow=length(lambdas), ncol=1)
for (i in 1:length(lambdas)) {
  P <- makeSurvivalMatrix(makeDeathRates, tstep, lambdas[i], a, w, d, .2, 5)
  B <- makeConstrainedBirthMatrix(flatBirthRates, tstep, lambdas[i], a, w, m, v, b, s)
  res<-eigen(t(P+B))
  df_lambda_r[i] <- log(Re(res$values[1]))
  x_pi <- res$vectors[,1] / sum(res$vectors[,1])
  # df_lambda[i,] <- c(Re(x_pi), rep(0, 5*w-lambdas[i]))
  df_lambda[i,] <- c(Re(x_pi), rep(NA, 5*w-lambdas[i]))
}
header <- paste('lambda', lambdas, sep=' ')
rownames(df_lambda) <- header
df_lambda <- tbl_df(t(df_lambda))
df_lambda <- df_lambda %>% mutate("age"=seq(1, 5*w))
df_lambda %>% gather(lambda, freq, -age) %>% ggplot() + aes(x=age,y=freq,color=lambda) + geom_line()

qplot(lambdas, df_lambda_r, geom='line')
```

### Alpha

Plot different values of alpha (first reproduction age) holding other things
constant.

```{r}
alphastep <- 1
alphas <- seq(1, (w-1), alphastep)
df_alpha <- matrix(0, nrow=length(alphas), ncol=lambda)
df_alpha_r <- matrix(0, nrow=length(alphas), ncol=1)
for (i in 1:length(alphas)) {
  P <- makeSurvivalMatrix(makeDeathRates, tstep, lambda, alphas[i], w, d, .2, 5)
  B <- makeConstrainedBirthMatrix(flatBirthRates, tstep, lambda, alphas[i], w, m, v, b, s)
  res<-eigen(t(P+B))
  df_alpha_r[i] <- log(Re(res$values[1]))
  x_pi <- res$vectors[,1] / sum(res$vectors[,1])
  df_alpha[i,] <- Re(x_pi)
}
header <- paste('alpha=0', alphas[1:ceiling(9/alphastep)], sep='')
header <- c(header, paste('alpha=', alphas[-(1:ceiling(9/alphastep))], sep=''))

rownames(df_alpha) <- header
colnames(df_alpha_r) <- 'r'
df_alpha <- tbl_df(t(df_alpha))

df_alpha <- df_alpha %>% mutate("age"=seq(1,lambda))
df_alpha %>%
  gather(alpha, freq, -age) %>%
  ggplot() + aes(x=age,y=freq,color=alpha) + geom_line()

df_alpha_r <- tbl_df(df_alpha_r) %>% mutate(alpha=alphas)
df_alpha_r %>%
  ggplot() + aes(x=alpha, y=r) + geom_line() + geom_hline(aes(yintercept=0))
```

### Omega

Plot different 'menopause' distribution (omega) holding other things constant.

```{r}
omegastep <- 2
omegas <- seq(a+1, lambda, omegastep)
df_omega <- matrix(0, nrow=length(omegas), ncol=lambda)
df_omega_r <- matrix(0, nrow=length(omegas), ncol=1)
for (i in 1:length(omegas)) {
  P <- makeSurvivalMatrix(makeDeathRates, tstep, lambda, a, w, d, .2, 5)
  B <- makeConstrainedBirthMatrix(flatBirthRates, tstep, lambda, a, omegas[i], m, v, b)
  res<-eigen(t(P+B))
  df_omega_r[i] <- log(Re(res$values[1]))
  x_pi <- res$vectors[,1] / sum(res$vectors[,1])
  df_omega[i,] <- Re(x_pi)
}
header <- paste('omega=', omegas, sep='')
rownames(df_omega) <- header
df_omega <- tbl_df(t(df_omega))

df_omega <- df_omega %>% mutate("age"=seq(1,lambda))
df_omega %>% gather(omega, freq, -age) %>% ggplot() + aes(x=age,y=freq,color=omega) + geom_line()

qplot(omegas, df_omega_r, geom='line')
```

### d (death)

Plot different death distribution holding other things constant.

```{r}
deaths <- seq(0, .45, .025)
df_death <- matrix(0, nrow=length(deaths), ncol=lambda)
df_death_r <- matrix(0, nrow=length(deaths), ncol=1)
for (i in 1:length(deaths)) {
  P <- makeSurvivalMatrix(makeDeathRates, tstep, lambda, a, w, deaths[i], .2, 5)
  B <- makeConstrainedBirthMatrix(flatBirthRates, tstep, lambda, a, w, m, v, b, s)
  res<-eigen(t(P+B))
  df_death_r[i] <- log(Re(res$values[1]))
  x_pi <- res$vectors[,1] / sum(res$vectors[,1])
  df_death[i,] <- Re(x_pi)
}
header <- paste('death', deaths, sep=' ')
rownames(df_death) <- header
rownames(df_death) <- header
df_death <- tbl_df(t(df_death))

df_death <- df_death %>% mutate("age"=seq(1,lambda))
df_death %>%
  gather(death, freq, -age) %>%
  ggplot() + aes(x=age,y=freq,color=death) + geom_line()

qplot(deaths, df_death_r, geom='line')
```


## High old mortality, flat birth rate

### Lambda

Plot different values of lambda (longevity) holding other things constant.

```{r}
lambdas <- seq(w, 5*w, w)
df_lambda <- matrix(0, nrow=length(lambdas), ncol=5*w)
df_lambda_r <- matrix(0, nrow=length(lambdas), ncol=1)
for (i in 1:length(lambdas)) {
  P <- makeSurvivalMatrix(makeDeathRates, tstep, lambdas[i], a, w, d, 5, .2)
  B <- makeConstrainedBirthMatrix(flatBirthRates, tstep, lambdas[i], a, w, m, v, b, s)
  res<-eigen(t(P+B))
  df_lambda_r[i] <- log(Re(res$values[1]))
  x_pi <- res$vectors[,1] / sum(res$vectors[,1])
  # df_lambda[i,] <- c(Re(x_pi), rep(0, 5*w-lambdas[i]))
  df_lambda[i,] <- c(Re(x_pi), rep(NA, 5*w-lambdas[i]))
}
header <- paste('lambda', lambdas, sep=' ')
rownames(df_lambda) <- header
df_lambda <- tbl_df(t(df_lambda))
df_lambda <- df_lambda %>% mutate("age"=seq(1, 5*w))
df_lambda %>%
  gather(lambda, freq, -age) %>%
  ggplot() + aes(x=age,y=freq,color=lambda) + geom_line()

qplot(lambdas, df_lambda_r, geom='line')
```

### Alpha

Plot different values of alpha (first reproduction age) holding other things
constant.

```{r}
alphastep <- 1
alphas <- seq(1, (w-1), alphastep)
df_alpha <- matrix(0, nrow=length(alphas), ncol=lambda)
df_alpha_r <- matrix(0, nrow=length(alphas), ncol=1)
for (i in 1:length(alphas)) {
  P <- makeSurvivalMatrix(makeDeathRates, tstep, lambda, alphas[i], w, d, 5, .2)
  B <- makeConstrainedBirthMatrix(flatBirthRates, tstep, lambda, alphas[i], w, m, v, b, s)
  res<-eigen(t(P+B))
  df_alpha_r[i] <- log(Re(res$values[1]))
  x_pi <- res$vectors[,1] / sum(res$vectors[,1])
  df_alpha[i,] <- Re(x_pi)
}
header <- paste('alpha=0', alphas[1:ceiling(9/alphastep)], sep='')
header <- c(header, paste('alpha=', alphas[-(1:ceiling(9/alphastep))], sep=''))

rownames(df_alpha) <- header
colnames(df_alpha_r) <- 'r'
df_alpha <- tbl_df(t(df_alpha))

df_alpha <- df_alpha %>% mutate("age"=seq(1,lambda))
df_alpha %>%
  gather(alpha, freq, -age) %>%
  ggplot() + aes(x=age,y=freq,color=alpha) + geom_line()

df_alpha_r <- tbl_df(df_alpha_r) %>% mutate(alpha=alphas)
df_alpha_r %>%
  ggplot() + aes(x=alpha, y=r) + geom_line() + geom_hline(aes(yintercept=0))
```

### Omega

Plot different 'menopause' distribution (omega) holding other things constant.

```{r}
omegastep <- 2
omegas <- seq(a+1, lambda, omegastep)
df_omega <- matrix(0, nrow=length(omegas), ncol=lambda)
df_omega_r <- matrix(0, nrow=length(omegas), ncol=1)
for (i in 1:length(omegas)) {
  P <- makeSurvivalMatrix(makeDeathRates, tstep, lambda, a, omegas[i], d, 5, .2)
  B <- makeConstrainedBirthMatrix(flatBirthRates, tstep, lambda, a, omegas[i], m, v, b, s)
  res<-eigen(t(P+B))
  df_omega_r[i] <- log(Re(res$values[1]))
  x_pi <- res$vectors[,1] / sum(res$vectors[,1])
  df_omega[i,] <- Re(x_pi)
}
header <- paste('omega=', omegas, sep='')
rownames(df_omega) <- header
df_omega <- tbl_df(t(df_omega))

df_omega <- df_omega %>% mutate("age"=seq(1,lambda))
df_omega %>%
  gather(omega, freq, -age) %>%
  ggplot() + aes(x=age,y=freq,color=omega) + geom_line()

qplot(omegas, df_omega_r, geom='line')
```

### d (death)

Plot different death distribution holding other things constant.

```{r}
deaths <- seq(0, .45, .025)
df_death <- matrix(0, nrow=length(deaths), ncol=lambda)
df_death_r <- matrix(0, nrow=length(deaths), ncol=1)
for (i in 1:length(deaths)) {
  P <- makeSurvivalMatrix(makeDeathRates, tstep, lambda, a, w, deaths[i], 5, .2)
  B <- makeConstrainedBirthMatrix(flatBirthRates, tstep, lambda, a, w, m, v, b, s)
  res<-eigen(t(P+B))
  df_death_r[i] <- log(Re(res$values[1]))
  x_pi <- res$vectors[,1] / sum(res$vectors[,1])
  df_death[i,] <- Re(x_pi)
}
header <- paste('death', deaths, sep=' ')
rownames(df_death) <- header
rownames(df_death) <- header
df_death <- tbl_df(t(df_death))

df_death <- df_death %>% mutate("age"=seq(1,lambda))
df_death %>%
  gather(death, freq, -age) %>%
  ggplot() + aes(x=age,y=freq,color=death) + geom_line()

qplot(deaths, df_death_r, geom='line')
```


# MCMC

Sample the population parameters to get estimates for maxima. (TERRIBLE, NO
GOOD, VERY BAD IDEA.)

I had been under the misimpression that the rate function could be constrained
such that the values converged to 0 at extreme values, but due to the log used
to compute the intrinsic growth rate, the function will inevitably diverge,
rendering the total volume in hyperspace infinite and the function
un-normalizable into a probability density function. Running a Metropolis
sampler on it is nonsense.

Sadly, the subject I chose for the project turned out to much, much more simple
than I thought and not very amenable to any of the more interesting topics we
learned. No EM to find demographic parameters given an empirical age
distribution, because there is no distribution for parameters. Numerical
methods are needed. No continuous MCMC, generator matrices, etc. because I
parameterized the chain such that it is occasionally transient and all states
move from one to another with each step.


## Uniform survival

```{r}
# Compute the eigenvalues for the transition matrix of the current population.
# Used for updating steps in the 'Metropolis' algorithm.
computeGrowthRate <- function(birthFxn, deathFxn, tstep, B, s, lambda, alpha,
                              omega, m, v, death, death_a, death_b) {
  P <- makeSurvivalMatrix(deathFxn, tstep, lambda, alpha, omega, d=death, death_a, death_b)
  B <- makeConstrainedBirthMatrix(birthFxn, tstep, lambda, alpha, omega, m, v, B, s)
  res<-eigen(t(P+B))
  return(log(Re(res$values[1])))
}

# Use to update all the demographic parameters and make sure they remain the
# correct data type (int or double) and remain in bounds.
updateParameter <- function(cur, move_m, move_sd, lower, upper, type) {
  if (type == 'int' ) {
    param <- cur + round(rnorm(1, mean=move_m, sd=move_sd)) # keep integer
  } else if (type == 'double') {
    param <- cur + rnorm(1, mean=move_m, sd=move_sd) # keep double
  }

  if (param > upper) {
    param <- updateParameter(cur, move_m, move_sd, lower, upper, type)
  } else if (param < lower) {
    param <- updateParameter(cur, move_m, move_sd, lower, upper, type)
  }
  # } else if (param == cur) {
  #   param <- updateParameter(cur, move_m, move_sd, lower, upper, type)
  # }

  return(param)
}

# Run a Markov chain to sample from the *not* distribution.
growthRateMCMC <- function(birthFxn, deathFxn, niter=1000, tstep=1, B0=20, s=1,
                           lambda0=50, alpha0=15, omega0=30, m0=20, v0=5,
                           death=.01, death_a=1, death_b=1) {
  d <- death
  d_a <- death_a
  d_b <- death_b

  lambdas <- rep(0, niter+1)
  alphas <- rep(0, niter+1)
  omegas <- rep(0, niter+1)
  ms <- rep(0, niter+1)
  vs <- rep(0, niter+1)
  Bs <- rep(0, niter+1)
  rs <- rep(0, niter+1)

  # Initialize
  lambdas[1] <- lambda0
  alphas[1] <- alpha0
  omegas[1] <- omega0
  ms[1] <- m0
  vs[1] <- v0
  Bs[1] <- B0
  rs[1] <- computeGrowthRate(birthFxn, deathFxn, tstep, Bs[1], s, lambdas[1],
                             alphas[1], omegas[1], ms[1], vs[1], d, d_a, d_b)

  # Update parameters
  for (i in 2:(niter+1)) {
    current_r <- rs[i-1]

    # Update alphas
    # print('update alpha')
    proposed_a <- updateParameter(alphas[i-1], 0, 1, 1, omegas[i-1]-1, 'int')
    new_r <- computeGrowthRate(birthFxn, deathFxn, tstep, Bs[i-1], s,
                               lambdas[i-1], proposed_a, omegas[i-1], ms[i-1],
                               vs[i-1], d, d_a, d_b)
    A <- new_r/current_r
    if (runif(1) < A) {
      alphas[i] <- proposed_a
    } else {
      alphas[i] <- alphas[i-1]
    }
    
    # Update omegas
    # print('update omega')
    proposed_w <- updateParameter(omegas[i-1], 0, 1, alphas[i]+1, lambdas[i-1]-1, 'int')
    new_r <- computeGrowthRate(birthFxn, deathFxn, tstep, Bs[i-1], s,
                               lambdas[i-1], alphas[i], proposed_w, ms[i-1],
                               vs[i-1], d, d_a, d_b)
    #A <- new_r/current_r
    A <- (new_r-current_r)/current_r
    if (runif(1) < A) {
      omegas[i] <- proposed_w
    } else {
      omegas[i] <- omegas[i-1]
    }
    
    # Update lambdas
    # print('update lambda')
    proposed_l <- updateParameter(lambdas[i-1], 0, 1, omegas[i], 10*lambda0, 'int')
    new_r <- computeGrowthRate(birthFxn, deathFxn, tstep, Bs[i-1], s,
                               proposed_l, alphas[i], omegas[i], ms[i-1],
                               vs[i-1], d, d_a, d_b)
    A <- new_r/current_r
    if (runif(1) < A) {
      lambdas[i] <- proposed_l
    } else {
      lambdas[i] <- lambdas[i-1]
    }
    
    # # Update means
    # proposed_m <- updateParameter(ms[i-1], 0, .5, 0, lambdas[i], 'double')
    # new_r <- computeGrowthRate(birthFxn, deathFxn, tstep, Bs[i-1], s,
    #                            lambdas[i], alphas[i], omegas[i], proposed_m,
    #                            vs[i-1], d, d_a, d_b)
    # A <- new_r/current_r
    # if (runif(1) < A) {
    #   ms[i] <- proposed_m
    # } else {
    #   ms[i] <- ms[i-1]
    # }
    # 
    # # Update variances
    # proposed_v <- updateParameter(vs[i-1], 0, .1, 0, lambdas[i], 'double')
    # new_r <- computeGrowthRate(birthFxn, deathFxn, tstep, Bs[i-1], s,
    #                            lambdas[i], alphas[i], omegas[i], ms[i],
    #                            proposed_v, d, d_a, d_b)
    # A <- new_r/current_r
    # if (runif(1) < A) {
    #   vs[i] <- proposed_v
    # } else {
    #   vs[i] <- vs[i-1]
    # }
    
    # Update litter size
    proposed_B <- updateParameter(Bs[i-1], 0, 1, 1, 1000, 'int')
    new_r <- computeGrowthRate(birthFxn, deathFxn, tstep, proposed_B, s,
                               lambdas[i], alphas[i], omegas[i], ms[i], vs[i],
                               d, d_a, d_b)
    A <- new_r/current_r
    if (runif(1) < A) {
      Bs[i] <- proposed_B
    } else {
      Bs[i] <- Bs[i-1]
    }

    rs[i] <- computeGrowthRate(birthFxn, deathFxn, tstep, Bs[i], s, lambdas[i],
                               alphas[i], omegas[i], ms[i], vs[i], d, d_a,
                               d_b)
    # print(rs[i])
    # print(lambdas[i])
    # print(alphas[i])
    # print(omegas[i])
    # print(Bs[i])
  }

  # return(list(r=rs, l=lambdas, a=alphas, w=omegas, m=ms, v=vs, B=Bs))
  return(list(r=rs, l=lambdas, a=alphas, w=omegas, B=Bs))
}
```

```{r}
niter <- 1000
tstep <- 1
B0 <- 10
s <- 1
lambda0 <- 50
alpha0 <- 15
omega0 <- 30
m0 <- 20
v0 <- 5
death <- .1
```

```{r}
death_a <- 1
death_b <- 1

res <- growthRateMCMC(flatBirthRates, makeDeathRates, niter, tstep, B0, s,
                      lambda0, alpha0, omega0, m0, v0, death, death_a, death_b)
df_res <- tbl_df(res)
```

```{r}
df_res[which.max(res$r),]
```

```{r}
ggplot(df_res) + aes(x=a, y=r) + geom_point()
ggplot(df_res) + aes(x=B, y=r) + geom_point()
ggplot(df_res) + aes(x=w, y=r) + geom_point()
ggplot(df_res) + aes(x=l, y=r) + geom_point()
```


### Early mortality

```{r}
# death_a <- .2
# death_b <- 5
# 
# res <- growthRateMCMC(flatBirthRates, makeDeathRates, niter, tstep, B0,
#                       lambda0, alpha0, omega0, m0, v0, death, death_a, death_b)
# df_res <- tbl_df(res)
```

```{r}
# df_res[which.max(res$r),]
```


### Late mortality

```{r}
# death_a <- 5
# death_b <- .2
# 
# res <- growthRateMCMC(flatBirthRates, makeDeathRates, niter, tstep, B0,
#                       lambda0, alpha0, omega0, m0, v0, death, death_a, death_b)
# df_res <- tbl_df(res)
```

```{r}
# df_res[which.max(res$r),]
```
