---
title: Project
author: Ansel George
output: pdf_document
---

```{r}
library(expm)
library(ggplot2)
library(dplyr)
library(tidyr)

set.seed(10)
```


# Functions

```{r}
# flatDeathRates <- function(t, d, a=1, b=1) {
#   return((1-d) * rep(1,length(t)))
# }

makeDeathRates <- function(t, a, w, d, d_a=1, d_b=1) {
  lambda <- max(t)+1
  p <- pbeta((t+1)/lambda, d_a, d_b) - pbeta(t/lambda, d_a, d_b)
  p <- 1-p
  p <- p/mean(p)
  p[a:w] <- p[a:w]*(1-1/t[a:w])
  p <- p * (1-d)
  # p[-(a:w)] <- p[-(a:w)]*(1-d)
  return(p)
}

makeDeathRates2 <- function(t, a, w, d, d_a=1, d_b=1) {
  lambda <- max(t)+1
  p <- pbeta((t+1)/lambda, d_a, d_b) - pbeta(t/lambda, d_a, d_b)
  p <- 1-p
  p <- p/mean(p)
  p <- p*(1-d)
  # p[a:w] <- p[a:w]*(1-2*d)
  # p[-(a:w)] <- p[-(a:w)]*(1-d)
  return(p)
}

makeSurvivalMatrix <- function(fxn, tstep, lambda, a, w, d=.1, d_a=1, d_b=1) {
  return(rbind(cbind(rep(0, lambda-1),
                     diag(lambda-1)),
               rep(0, lambda)) * fxn(seq(0, (lambda-1)*tstep), a, w, d, d_a, d_b))
}

# makeSurvivalMatrix <- function(fxn, tstep, lambda) {
#   d <- fxn(seq(0, (lambda)*tstep, tstep))
#   Q <- matrix(0, nrow=(lambda+1), ncol=(lambda+1))
#   Q <- -diag(lambda+1) + rbind(cbind(rep(0, lambda), diag(lambda)), rep(0, lambda+1)) %*% diag(1-d)
#   Q[,(lambda+1)] <- d
#   Q[(lambda+1),(lambda+1)] <- 0
#   Q[(lambda),(lambda+1)] <- 1
# 
#   return(expm(Q)[-(lambda+1),-(lambda+1)])
# }

flatBirthRates <- function(t, a, w, m, v) {
  b <- rep(0, length(t))
  b[a:w] <- 1
  return(b)
}

normalBirthRates <- function(age, a, w, m, v) {
  p <- dnorm(age, m, sqrt(v))
  return((w-a+1)*p/sum(p))
}

makeBirthMatrix <- function(fxn, tstep, lambda, a=0, w=0, m=1, v=0, b=1) {
  B <- matrix(0, nrow=lambda, ncol=lambda)
  B[1,] <- b*fxn(seq(0, (lambda-1)*tstep, tstep), a, w, m, v)
  return(t(B))
}

makeConstrainedBirthMatrix <- function(fxn, tstep, lambda, a=0, w=0, m=1, v=0, b=1) {
  B <- matrix(0, nrow=lambda, ncol=lambda)
  # B_survive <- diag(pmin(1, (log(seq(0,lambda-1)+1)/b)^2))
  # B_survive <- diag(pmin(1, (log(seq(0,lambda-1)+1)/b)))
  B_survive <- diag(pmin(1, (seq(0,lambda-1)/b^2)))
  B[1,] <- b*fxn(seq(0, (lambda-1)*tstep, tstep), a, w, m, v) %*% B_survive
  return(t(B))
}

makeConstrainedBirthMatrix2 <- function(fxn, tstep, lambda, a=0, w=0, m=1, v=0, b=1, s=2) {
  B <- matrix(0, nrow=lambda, ncol=lambda)
  # B_survive <- diag(pmin(1, (seq(0,lambda-1)/b)^s))
  B_survive <- diag((rep(1, lambda)/b)^s)
  B[1,] <- b*fxn(seq(0, (lambda-1)*tstep, tstep), a, w, m, v) %*% B_survive
  return(t(B))
}

simulateGenerations <- function(x, P, B, n) {
  x_new <- x %*% ((P+B) %^% n)
  x_new <- x_new / sum(x_new)
  return(x_new)
  
  # X <- matrix(0, nrow=n, ncol=length(x))
  # N <- matrix(0, nrow=n, ncol=1)
  # X[1,] <- x
  # N[1] <- sum(x)
  # 
  # for (i in 2:n) {
  #   tmp <- X[i-1,] %*% (P+B)
  #   X[i,] <- tmp / sum(tmp)
  #   N[i] <- sum(tmp)
  # }
  # return(X)
}
```


# Global variables

```{r}
tstep <- 1
# a <- 15
a <- 20
w <- 30
b <- 5
d <- .1
m <- 20
v <- 5
lambda <- 50

x <- rep(1/lambda, lambda)
```


# Simple simulations

## Flat birth/death rates

```{r}
P <- makeSurvivalMatrix(makeDeathRates2, tstep, lambda, a, w, d, 1, 1)
# B <- makeConstrainedBirthMatrix(flatBirthRates, tstep, lambda, a, w, m, v, b)
B <- makeBirthMatrix(flatBirthRates, tstep, lambda, a, w, m, v, b)

res<-eigen(t(P+B))
log(res$values[1])
x_pi <- res$vectors[,1] / sum(res$vectors[,1])

age <- seq(1, lambda, tstep)
qplot(age, x)
qplot(age, Re(x_pi))

# log(x_pi[2])-log(x_pi[1])
# log(x_pi[7])-log(x_pi[6])
```

Plot different values of lambda (longevity) holding other things constant.

```{r}
lambdas <- seq(w, 5*w, w)
df_lambda <- matrix(0, nrow=length(lambdas), ncol=5*w)
df_lambda_r <- matrix(0, nrow=length(lambdas), ncol=1)
for (i in 1:length(lambdas)) {
  # P <- makeSurvivalMatrix(makeDeathRates, tstep, lambdas[i], a, w, d, 1, 1) 
  P <- makeSurvivalMatrix(makeDeathRates2, tstep, lambdas[i], a, w, d, 1, 1) 
  # B <- makeBirthMatrix(flatBirthRates, tstep, lambdas[i], a, w, m, v, b)
  # B <- makeConstrainedBirthMatrix(flatBirthRates, tstep, lambdas[i], a, w, m, v, b)
  B <- makeConstrainedBirthMatrix2(flatBirthRates, tstep, lambdas[i], a, w, m, v, b, 2)
  res<-eigen(t(P+B))
  df_lambda_r[i] <- log(Re(res$values[1]))
  x_pi <- res$vectors[,1] / sum(res$vectors[,1])
  # df_lambda[i,] <- c(Re(x_pi), rep(0, 5*w-lambdas[i]))
  df_lambda[i,] <- c(Re(x_pi), rep(NA, 5*w-lambdas[i]))
}
header <- paste('lambda', lambdas, sep=' ')
rownames(df_lambda) <- header
df_lambda <- tbl_df(t(df_lambda))
df_lambda <- df_lambda %>% mutate("age"=seq(1, 5*w))
f1 <- df_lambda %>% gather(lambda, freq, -age) %>% ggplot() + aes(x=age,y=freq,color=lambda) + geom_line()
ggsave(file="freq_lambda-1-1.svg", plot=f1)

f2 <- qplot(lambdas, df_lambda_r, geom='line')
ggsave(file="r_lambda-1-1.svg", plot=f2)
```

Plot different values of alpha (first reproduction age) holding other things
constant.

```{r}
alphastep <- 1
alphas <- seq(1, (w-1), alphastep)
df_alpha <- matrix(0, nrow=length(alphas), ncol=lambda)
df_alpha_r <- matrix(0, nrow=length(alphas), ncol=1)
for (i in 1:length(alphas)) {
  # P <- makeSurvivalMatrix(makeDeathRates, tstep, lambda, alphas[i], w, d, 1, 1)
  P <- makeSurvivalMatrix(makeDeathRates2, tstep, lambda, alphas[i], w, d, 1, 1)
  # B <- makeBirthMatrix(flatBirthRates, tstep, lambda, alphas[i], w, m, v, b)
  # B <- makeConstrainedBirthMatrix(flatBirthRates, tstep, lambda, alphas[i], w, m, v, b)
  B <- makeConstrainedBirthMatrix2(flatBirthRates, tstep, lambda, alphas[i], w, m, v, b, 1)
  res<-eigen(t(P+B))
  df_alpha_r[i] <- log(Re(res$values[1]))
  x_pi <- res$vectors[,1] / sum(res$vectors[,1])
  df_alpha[i,] <- Re(x_pi)
}
header <- paste('alpha=0', alphas[1:ceiling(9/alphastep)], sep='')
header <- c(header, paste('alpha=', alphas[-(1:ceiling(9/alphastep))], sep=''))

rownames(df_alpha) <- header
colnames(df_alpha_r) <- 'r'
df_alpha <- tbl_df(t(df_alpha))

# df_alpha <- df_alpha %>% mutate("age"=seq(1,lambda))
# f1 <- df_alpha %>% gather(alpha, freq, -age) %>% ggplot() + aes(x=age,y=freq,color=alpha) + geom_line()
# ggsave(file="freq_alpha-1-1.svg", plot=f1, width=6, height=4)

# df_alpha_r <- tbl_df(df_alpha_r) %>% mutate(alpha=alphas)
# f2 <- df_alpha_r %>% ggplot() + aes(x=alpha, y=r) + geom_line()
# ggsave(file="r_alpha-1-1.svg", plot=f2, width=6, height=4)
```

```{r}
alphastep <- 1
alphas <- seq(1, (w-1), alphastep)
df_alpha <- matrix(0, nrow=length(alphas), ncol=lambda)
df_alpha_r2 <- matrix(0, nrow=length(alphas), ncol=1)
for (i in 1:length(alphas)) {
  # P <- makeSurvivalMatrix(makeDeathRates, tstep, lambda, alphas[i], w, d, 1, 1)
  P <- makeSurvivalMatrix(makeDeathRates2, tstep, lambda, alphas[i], w, d, 1, 1)
  # B <- makeBirthMatrix(flatBirthRates, tstep, lambda, alphas[i], w, m, v, b)
  # B <- makeConstrainedBirthMatrix(flatBirthRates, tstep, lambda, alphas[i], w, m, v, b)
  B <- makeConstrainedBirthMatrix2(flatBirthRates, tstep, lambda, alphas[i], w, m, v, b, 2)
  res<-eigen(t(P+B))
  df_alpha_r2[i] <- log(Re(res$values[1]))
  x_pi <- res$vectors[,1] / sum(res$vectors[,1])
  df_alpha[i,] <- Re(x_pi)
}
header <- paste('alpha=0', alphas[1:ceiling(9/alphastep)], sep='')
header <- c(header, paste('alpha=', alphas[-(1:ceiling(9/alphastep))], sep=''))

rownames(df_alpha) <- header
colnames(df_alpha_r2) <- 'r'
df_alpha <- tbl_df(t(df_alpha))

# df_alpha <- df_alpha %>% mutate("age"=seq(1,lambda))
# f1 <- df_alpha %>% gather(alpha, freq, -age) %>% ggplot() + aes(x=age,y=freq,color=alpha) + geom_line()
# ggsave(file="freq_alpha-1-1.svg", plot=f1, width=6, height=4)
# 
# df_alpha_r <- tbl_df(df_alpha_r) %>% mutate(alpha=alphas)
# f2 <- df_alpha_r %>% ggplot() + aes(x=alpha, y=r) + geom_line()
# ggsave(file="r_alpha-1-1.svg", plot=f2, width=6, height=4)
```

```{r}
alphastep <- 1
alphas <- seq(1, (w-1), alphastep)
df_alpha <- matrix(0, nrow=length(alphas), ncol=lambda)
df_alpha_r3 <- matrix(0, nrow=length(alphas), ncol=1)
for (i in 1:length(alphas)) {
  # P <- makeSurvivalMatrix(makeDeathRates, tstep, lambda, alphas[i], w, d, 1, 1)
  P <- makeSurvivalMatrix(makeDeathRates2, tstep, lambda, alphas[i], w, d, 1, 1)
  # B <- makeBirthMatrix(flatBirthRates, tstep, lambda, alphas[i], w, m, v, b)
  # B <- makeConstrainedBirthMatrix(flatBirthRates, tstep, lambda, alphas[i], w, m, v, b)
  B <- makeConstrainedBirthMatrix2(flatBirthRates, tstep, lambda, alphas[i], w, m, v, b, .5)
  res<-eigen(t(P+B))
  df_alpha_r3[i] <- log(Re(res$values[1]))
  x_pi <- res$vectors[,1] / sum(res$vectors[,1])
  df_alpha[i,] <- Re(x_pi)
}
header <- paste('alpha=0', alphas[1:ceiling(9/alphastep)], sep='')
header <- c(header, paste('alpha=', alphas[-(1:ceiling(9/alphastep))], sep=''))

rownames(df_alpha) <- header
colnames(df_alpha_r3) <- 'r'
df_alpha <- tbl_df(t(df_alpha))

# df_alpha <- df_alpha %>% mutate("age"=seq(1,lambda))
# f1 <- df_alpha %>% gather(alpha, freq, -age) %>% ggplot() + aes(x=age,y=freq,color=alpha) + geom_line()
# ggsave(file="freq_alpha-1-1.svg", plot=f1, width=6, height=4)
# 
# df_alpha_r <- tbl_df(df_alpha_r) %>% mutate(alpha=alphas)
# f2 <- df_alpha_r %>% ggplot() + aes(x=alpha, y=r) + geom_line()
# ggsave(file="r_alpha-1-1.svg", plot=f2, width=6, height=4)
```

Plot different 'menopause' distribution (omega) holding other things constant.

```{r}
omegastep <- 2
omegas <- seq(a+1, lambda, omegastep)
df_omega <- matrix(0, nrow=length(omegas), ncol=lambda)
df_omega_r <- matrix(0, nrow=length(omegas), ncol=1)
for (i in 1:length(omegas)) {
  P <- makeSurvivalMatrix(makeDeathRates, tstep, lambda, a, w, d, 1, 1)
  # B <- makeBirthMatrix(flatBirthRates, tstep, lambda, a, omegas[i], m, v, b)
  # B <- makeConstrainedBirthMatrix(flatBirthRates, tstep, lambda, a, omegas[i], m, v, b)
  B <- makeConstrainedBirthMatrix2(flatBirthRates, tstep, lambda, a, omegas[i], m, v, b, 1)
  res<-eigen(t(P+B))
  df_omega_r[i] <- log(Re(res$values[1]))
  x_pi <- res$vectors[,1] / sum(res$vectors[,1])
  df_omega[i,] <- Re(x_pi)
}
df_omega_r2 <- matrix(0, nrow=length(omegas), ncol=1)
for (i in 1:length(omegas)) {
  P <- makeSurvivalMatrix(makeDeathRates, tstep, lambda, a, w, d, 1, 1)
  # B <- makeBirthMatrix(flatBirthRates, tstep, lambda, a, omegas[i], m, v, b)
  # B <- makeConstrainedBirthMatrix(flatBirthRates, tstep, lambda, a, omegas[i], m, v, b)
  B <- makeConstrainedBirthMatrix2(flatBirthRates, tstep, lambda, a, omegas[i], m, v, b, 2)
  res<-eigen(t(P+B))
  df_omega_r2[i] <- log(Re(res$values[1]))
}
df_omega_r3 <- matrix(0, nrow=length(omegas), ncol=1)
for (i in 1:length(omegas)) {
  P <- makeSurvivalMatrix(makeDeathRates, tstep, lambda, a, w, d, 1, 1)
  # B <- makeBirthMatrix(flatBirthRates, tstep, lambda, a, omegas[i], m, v, b)
  # B <- makeConstrainedBirthMatrix(flatBirthRates, tstep, lambda, a, omegas[i], m, v, b)
  B <- makeConstrainedBirthMatrix2(flatBirthRates, tstep, lambda, a, omegas[i], m, v, b, 0.5)
  res<-eigen(t(P+B))
  df_omega_r3[i] <- log(Re(res$values[1]))
}
header <- paste('omega=', omegas, sep='')
rownames(df_omega) <- header
df_omega <- tbl_df(t(df_omega))

df_omega <- df_omega %>% mutate("age"=seq(1,lambda))
f1 <- df_omega %>% gather(omega, freq, -age) %>% ggplot() + aes(x=age,y=freq,color=omega) + geom_line()
ggsave(file="freq_omega-1-1.svg", plot=f1, width=6, height=4)

# f2 <- qplot(omegas, df_omega_r, geom='line', ylab='r', xlab='age at end of fertility')
# ggsave(file="r_omega-1-1.svg", plot=f2, width=6, height=4)
tmp<-data.frame(df_omega_r, df_omega_r2, df_omega_r3)
colnames(tmp) <- c('s=1', 's=2', 's=0.5')
df_tmp <- tbl_df(tmp) %>% mutate('omega'=omegas)
f <- df_tmp %>% gather(s, r, -omega) %>% ggplot() + aes(x=omega, y=r, color=s) + geom_line()
ggsave(file="r_omega-1-1_extra.svg", plot=f, width=6, height=4)
```

Plot different 'litter size' distribution (b) holding other things constant.

```{r}
bstep <- 1
bs <- seq(1, 10, bstep)
df_b <- matrix(0, nrow=length(bs), ncol=lambda)
df_b_r <- matrix(0, nrow=length(bs), ncol=1)
for (i in 1:length(bs)) {
  # P <- makeSurvivalMatrix(makeDeathRates, tstep, lambda, a, w, d, 1, 1)
  P <- makeSurvivalMatrix(makeDeathRates2, tstep, lambda, a, w, d, 1, 1)
  # B <- makeBirthMatrix(flatBirthRates, tstep, lambda, a, w, m, v, bs[i])
  # B <- makeConstrainedBirthMatrix(flatBirthRates, tstep, lambda, a, bs[i], m, v, b)
  B <- makeConstrainedBirthMatrix2(flatBirthRates, tstep, lambda, a, w, m, v, bs[i], 1)
  res<-eigen(t(P+B))
  df_b_r[i] <- log(Re(res$values[1]))
  x_pi <- res$vectors[,1] / sum(res$vectors[,1])
  df_b[i,] <- Re(x_pi)
}
df_b_r2 <- matrix(0, nrow=length(bs), ncol=1)
for (i in 1:length(bs)) {
  # P <- makeSurvivalMatrix(makeDeathRates, tstep, lambda, a, w, d, 1, 1)
  P <- makeSurvivalMatrix(makeDeathRates2, tstep, lambda, a, w, d, 1, 1)
  # B <- makeBirthMatrix(flatBirthRates, tstep, lambda, a, w, m, v, bs[i])
  # B <- makeConstrainedBirthMatrix(flatBirthRates, tstep, lambda, a, bs[i], m, v, b)
  B <- makeConstrainedBirthMatrix2(flatBirthRates, tstep, lambda, a, w, m, v, bs[i], 2)
  res<-eigen(t(P+B))
  df_b_r2[i] <- log(Re(res$values[1]))
}
df_b_r3 <- matrix(0, nrow=length(bs), ncol=1)
for (i in 1:length(bs)) {
  # P <- makeSurvivalMatrix(makeDeathRates, tstep, lambda, a, w, d, 1, 1)
  P <- makeSurvivalMatrix(makeDeathRates2, tstep, lambda, a, w, d, 1, 1)
  # B <- makeBirthMatrix(flatBirthRates, tstep, lambda, a, w, m, v, bs[i])
  # B <- makeConstrainedBirthMatrix(flatBirthRates, tstep, lambda, a, bs[i], m, v, b)
  B <- makeConstrainedBirthMatrix2(flatBirthRates, tstep, lambda, a, w, m, v, bs[i], .5)
  res<-eigen(t(P+B))
  df_b_r3[i] <- log(Re(res$values[1]))
}
header <- paste('b=', bs, sep='')
rownames(df_b) <- header
df_b <- tbl_df(t(df_b))

df_b <- df_b %>% mutate("age"=seq(1,lambda))
f1 <- df_b %>% gather(b, freq, -age) %>% ggplot() + aes(x=age,y=freq,color=b) + geom_line()
ggsave(file="freq_b-1-1.svg", plot=f1, width=6, height=4)

# f2 <- qplot(bs, df_b_r, geom='line', xlab='litter size', ylab='r')
# ggsave(file="r_b-1-1.svg", plot=f2, width=6, height=4)
tmp<-data.frame(df_b_r, df_b_r2, df_b_r3)
colnames(tmp) <- c('s=1', 's=2', 's=0.5')
df_tmp <- tbl_df(tmp) %>% mutate('b'=bs)
f <- df_tmp %>% gather(s, r, -b) %>% ggplot() + aes(x=b, y=r, color=s) + geom_line()
ggsave(file="r_b-1-1_extra.svg", plot=f, width=6, height=4)
```

Plot different 'strength of constraint' (s) holding other things constant.

```{r}
sstep <- .1
ss <- seq(0, 5, sstep)
df_s <- matrix(0, nrow=length(ss), ncol=lambda)
df_s_r <- matrix(0, nrow=length(ss), ncol=1)
for (i in 1:length(ss)) {
  # P <- makeSurvivalMatrix(makeDeathRates, tstep, lambda, a, w, d, 1, 1)
  P <- makeSurvivalMatrix(makeDeathRates2, tstep, lambda, a, w, d, 1, 1)
  # B <- makeBirthMatrix(flatBirthRates, tstep, lambda, a, w, m, v, bs[i])
  # B <- makeConstrainedBirthMatrix(flatBirthRates, tstep, lambda, a, bs[i], m, v, b)
  B <- makeConstrainedBirthMatrix2(flatBirthRates, tstep, lambda, a, w, m, v, b, ss[i])
  res<-eigen(t(P+B))
  df_s_r[i] <- log(Re(res$values[1]))
  x_pi <- res$vectors[,1] / sum(res$vectors[,1])
  df_s[i,] <- Re(x_pi)
}
df_s_r2 <- matrix(0, nrow=length(ss), ncol=1)
for (i in 1:length(ss)) {
  # P <- makeSurvivalMatrix(makeDeathRates, tstep, lambda, a, w, d, 1, 1)
  P <- makeSurvivalMatrix(makeDeathRates2, tstep, lambda, a, w, d, 1, 1)
  # B <- makeBirthMatrix(flatBirthRates, tstep, lambda, a, w, m, v, bs[i])
  # B <- makeConstrainedBirthMatrix(flatBirthRates, tstep, lambda, a, bs[i], m, v, b)
  B <- makeConstrainedBirthMatrix2(flatBirthRates, tstep, lambda, a, w, m, v, 1, ss[i])
  res<-eigen(t(P+B))
  df_s_r2[i] <- log(Re(res$values[1]))
}
df_s_r3 <- matrix(0, nrow=length(ss), ncol=1)
for (i in 1:length(ss)) {
  # P <- makeSurvivalMatrix(makeDeathRates, tstep, lambda, a, w, d, 1, 1)
  P <- makeSurvivalMatrix(makeDeathRates2, tstep, lambda, a, w, d, 1, 1)
  # B <- makeBirthMatrix(flatBirthRates, tstep, lambda, a, w, m, v, bs[i])
  # B <- makeConstrainedBirthMatrix(flatBirthRates, tstep, lambda, a, bs[i], m, v, b)
  B <- makeConstrainedBirthMatrix2(flatBirthRates, tstep, lambda, a, w, m, v, 10, ss[i])
  res<-eigen(t(P+B))
  df_s_r3[i] <- log(Re(res$values[1]))
}
header <- paste('s=', ss, sep='')
rownames(df_s) <- header
df_s <- tbl_df(t(df_s))

df_s <- df_s %>% mutate("age"=seq(1,lambda))
f1 <- df_s %>% gather(s, freq, -age) %>% ggplot() + aes(x=age,y=freq,color=s) + geom_line()
ggsave(file="freq_s-1-1.svg", plot=f1, width=6, height=4)

# f2 <- qplot(ss, df_s_r, geom='line', xlab='strength of constraint', ylab='r')
# ggsave(file="r_s-1-1.svg", plot=f2, width=6, height=4)
tmp<-data.frame(df_s_r, df_s_r2, df_s_r3)
colnames(tmp) <- c('b=5', 'b=1', 'b=10')
df_tmp <- tbl_df(tmp) %>% mutate('s'=ss)
f <- df_tmp %>% gather(b, r, -s) %>% ggplot() + aes(x=s, y=r, color=b) + geom_line()
ggsave(file="r_s-1-1_extra.svg", plot=f, width=6, height=4)
```



Plot different death distribution holding other things constant.

```{r}
deaths <- seq(0, .45, .025)
df_death <- matrix(0, nrow=length(deaths), ncol=lambda)
df_death_r <- matrix(0, nrow=length(deaths), ncol=1)
for (i in 1:length(deaths)) {
  P <- makeSurvivalMatrix(makeDeathRates, tstep, lambda, a, w, deaths[i], 1, 1)
  # B <- makeBirthMatrix(flatBirthRates, tstep, lambda, a, w, m, v, b)
  B <- makeConstrainedBirthMatrix(flatBirthRates, tstep, lambda, a, w, m, v, b)
  res<-eigen(t(P+B))
  df_death_r[i] <- log(Re(res$values[1]))
  x_pi <- res$vectors[,1] / sum(res$vectors[,1])
  df_death[i,] <- Re(x_pi)
}
header <- paste('death', deaths, sep=' ')
rownames(df_death) <- header
rownames(df_death) <- header
df_death <- tbl_df(t(df_death))

df_death <- df_death %>% mutate("age"=seq(1,lambda))
df_death %>% gather(death, freq, -age) %>% ggplot() + aes(x=age,y=freq,color=death) + geom_point()

qplot(deaths, df_death_r, geom='line')
```

### Use EM to estimate parameters

```{r}

```


```{r}
f <- df_alpha$`alpha=01`
age <- seq(0, lambda-1)
qplot(age, f)
```


## Normal birth, flat death

Plot different values of lambda (longevity) holding other things constant.

```{r}
# lambdas <- seq(m, 3*m, 2)
# t_lambda <- matrix(0, nrow=length(lambdas), ncol=3*m)
# t_lambda_r <- matrix(0, nrow=length(lambdas), ncol=1)
# for (i in 1:length(lambdas)) {
#   P <- makeSurvivalMatrix(makeDeathRates, tstep, lambdas[i], d, 1, 1)
#   B <- makeBirthMatrix(normalBirthRates, tstep, lambdas[i], a, w, m, v, b)
#   res<-eigen(t(P+B))
#   t_lambda_r[i] <- log(Re(res$values[1]))
#   x_pi <- res$vectors[,1] / sum(res$vectors[,1])
#   t_lambda[i,] <- c(Re(x_pi), rep(NA, 3*m-lambdas[i]))
# }
# header <- paste('lambda', lambdas, sep=' ')
# rownames(t_lambda) <- header
# df_lambda <- tbl_df(t(t_lambda))
# df_lambda <- df_lambda %>% mutate("age"=seq(1, 3*m))
# df_lambda %>% gather(lambda, freq, -age) %>% ggplot() + aes(x=age,y=freq,color=lambda) + geom_line()
# 
# qplot(lambdas, t_lambda_r, geom='line')
# qplot(x=seq(1, length(B[,1])), B[,1])
```

Plot different values of mean reproduction age (m) holding other things
constant.

```{r}
# meanstep <- 1
# means <- seq(ceiling(m/2), 2*m, meanstep)
# df_mean <- matrix(0, nrow=length(means), ncol=lambda)
# df_mean_r <- matrix(0, nrow=length(means), ncol=1)
# for (i in 1:length(means)) {
#   P <- makeSurvivalMatrix(makeDeathRates, tstep, lambda, d, 1, 1)
#   B <- makeBirthMatrix(normalBirthRates, tstep, lambda, a, w, means[i], v, b)
#   res<-eigen(t(P+B))
#   df_mean_r[i] <- log(Re(res$values[1]))
#   x_pi <- res$vectors[,1] / sum(res$vectors[,1])
#   df_mean[i,] <- Re(x_pi)
# }
# header <- paste('m=', means, sep='')
# 
# rownames(df_mean) <- header
# colnames(df_mean_r) <- 'r'
# df_mean <- tbl_df(t(df_mean))
# 
# df_mean <- df_mean %>% mutate("age"=seq(1,lambda))
# df_mean %>% gather(m, freq, -age) %>% ggplot() + aes(x=age,y=freq,color=m) + geom_line()
# 
# df_mean_r <- tbl_df(df_mean_r) %>% mutate(m=means)
# df_mean_r %>% ggplot() + aes(x=m, y=r) + geom_line() + geom_hline(aes(yintercept=0), color='red')
```

Plot variance of different distributions other things constant.

```{r}
# vstep <- 1
# vs <- seq(2, 10)
# t_v <- matrix(0, nrow=length(vs), ncol=lambda)
# t_v_r <- matrix(0, nrow=length(vs), ncol=1)
# for (i in 1:length(vs)) {
#   P <- makeSurvivalMatrix(makeDeathRates, tstep, lambda, d, 1, 1)
#   B <- makeBirthMatrix(normalBirthRates, tstep, lambda, a, w, m, vs[i], b)
#   res <- eigen(t(P+B))
#   t_v_r[i] <- log(Re(res$values[1]))
#   x_pi <- res$vectors[,1] / sum(res$vectors[,1])
#   t_v[i,] <- Re(x_pi)
# }
# header <- paste('v=0', vs[1:8], sep='')
# header <- c(header, paste('v=', vs[-(1:8)], sep=''))
# 
# rownames(t_v) <- header
# colnames(t_v_r) <- 'r'
# df_v <- tbl_df(t(t_v))
# 
# df_v <- df_v %>% mutate("age"=seq(1,lambda))
# df_v %>% gather(v, freq, -age) %>% ggplot() + aes(x=age,y=freq,color=v) + geom_line()
# 
# df_v_r <- tbl_df(t_v_r) %>% mutate(v=vs)
# df_v_r %>% ggplot() + aes(x=v, y=r) + geom_line() + geom_hline(aes(yintercept=0))
```

Plot different death distribution holding other things constant.

```{r}
# deaths <- seq(0, .95, .05)
# df_death <- matrix(0, nrow=length(deaths), ncol=lambda)
# df_death_r <- matrix(0, nrow=length(deaths), ncol=1)
# for (i in 1:length(deaths)) {
#   P <- makeSurvivalMatrix(makeDeathRates, tstep, lambda, deaths[i], 1, 1)
#   B <- makeBirthMatrix(normalBirthRates, tstep, lambda, a, w, m, v, b)
#   res<-eigen(t(P+B))
#   df_death_r[i] <- log(Re(res$values[1]))
#   x_pi <- res$vectors[,1] / sum(res$vectors[,1])
#   df_death[i,] <- Re(x_pi)
# }
# header <- paste('death=', deaths, sep='')
# rownames(df_death) <- header
# rownames(df_death) <- header
# df_death <- tbl_df(t(df_death))
# 
# df_death <- df_death %>% mutate("age"=seq(1,lambda))
# df_death %>% gather(death, freq, -age) %>% ggplot() + aes(x=age,y=freq,color=death) + geom_line()
# 
# qplot(deaths, df_death_r, geom='line')
```

## Add skewed survivorship function

### High young mortality, flat birth rate 

Plot different values of lambda (longevity) holding other things constant.

```{r}
lambdas <- seq(w, 5*w, w)
df_lambda <- matrix(0, nrow=length(lambdas), ncol=5*w)
df_lambda_r <- matrix(0, nrow=length(lambdas), ncol=1)
for (i in 1:length(lambdas)) {
  P <- makeSurvivalMatrix(makeDeathRates, tstep, lambdas[i], a, w, d, .2, 5)
  # B <- makeBirthMatrix(flatBirthRates, tstep, lambdas[i], a, w, m, v, b)
  B <- makeConstrainedBirthMatrix(flatBirthRates, tstep, lambdas[i], a, w, m, v, b)
  res<-eigen(t(P+B))
  df_lambda_r[i] <- log(Re(res$values[1]))
  x_pi <- res$vectors[,1] / sum(res$vectors[,1])
  # df_lambda[i,] <- c(Re(x_pi), rep(0, 5*w-lambdas[i]))
  df_lambda[i,] <- c(Re(x_pi), rep(NA, 5*w-lambdas[i]))
}
header <- paste('lambda', lambdas, sep=' ')
rownames(df_lambda) <- header
df_lambda <- tbl_df(t(df_lambda))
df_lambda <- df_lambda %>% mutate("age"=seq(1, 5*w))
df_lambda %>% gather(lambda, freq, -age) %>% ggplot() + aes(x=age,y=freq,color=lambda) + geom_line()

qplot(lambdas, df_lambda_r, geom='line')
```

Plot different values of alpha (first reproduction age) holding other things
constant.

```{r}
alphastep <- 1
alphas <- seq(1, (w-1), alphastep)
df_alpha <- matrix(0, nrow=length(alphas), ncol=lambda)
df_alpha_r <- matrix(0, nrow=length(alphas), ncol=1)
for (i in 1:length(alphas)) {
  # P <- makeSurvivalMatrix(makeDeathRates, tstep, lambda, alphas[i], w, d, .2, 5)
  P <- makeSurvivalMatrix(makeDeathRates2, tstep, lambda, alphas[i], w, d, .2, 5)
  B <- makeBirthMatrix(flatBirthRates, tstep, lambda, alphas[i], w, m, v, b)
  # B <- makeConstrainedBirthMatrix(flatBirthRates, tstep, lambda, alphas[i], w, m, v, b)
  res<-eigen(t(P+B))
  df_alpha_r[i] <- log(Re(res$values[1]))
  x_pi <- res$vectors[,1] / sum(res$vectors[,1])
  df_alpha[i,] <- Re(x_pi)
}
header <- paste('alpha=0', alphas[1:ceiling(9/alphastep)], sep='')
header <- c(header, paste('alpha=', alphas[-(1:ceiling(9/alphastep))], sep=''))

rownames(df_alpha) <- header
colnames(df_alpha_r) <- 'r'
df_alpha <- tbl_df(t(df_alpha))

df_alpha <- df_alpha %>% mutate("age"=seq(1,lambda))
df_alpha %>% gather(alpha, freq, -age) %>% ggplot() + aes(x=age,y=freq,color=alpha) + geom_line()

df_alpha_r <- tbl_df(df_alpha_r) %>% mutate(alpha=alphas)
df_alpha_r %>% ggplot() + aes(x=alpha, y=r) + geom_line() + geom_hline(aes(yintercept=0))
```

Plot different 'menopause' distribution (omega) holding other things constant.

```{r}
omegastep <- 2
omegas <- seq(a+1, lambda, omegastep)
df_omega <- matrix(0, nrow=length(omegas), ncol=lambda)
df_omega_r <- matrix(0, nrow=length(omegas), ncol=1)
for (i in 1:length(omegas)) {
  P <- makeSurvivalMatrix(makeDeathRates, tstep, lambda, a, w, d, .2, 5)
  B <- makeConstrainedBirthMatrix(flatBirthRates, tstep, lambda, a, omegas[i], m, v, b)
  res<-eigen(t(P+B))
  df_omega_r[i] <- log(Re(res$values[1]))
  x_pi <- res$vectors[,1] / sum(res$vectors[,1])
  df_omega[i,] <- Re(x_pi)
}
header <- paste('omega=', omegas, sep='')
rownames(df_omega) <- header
df_omega <- tbl_df(t(df_omega))

df_omega <- df_omega %>% mutate("age"=seq(1,lambda))
df_omega %>% gather(omega, freq, -age) %>% ggplot() + aes(x=age,y=freq,color=omega) + geom_line()

qplot(omegas, df_omega_r, geom='line')
```

Plot different death distribution holding other things constant.

```{r}
deaths <- seq(0, .45, .025)
df_death <- matrix(0, nrow=length(deaths), ncol=lambda)
df_death_r <- matrix(0, nrow=length(deaths), ncol=1)
for (i in 1:length(deaths)) {
  P <- makeSurvivalMatrix(makeDeathRates, tstep, lambda, a, w, deaths[i], .2, 5)
  # B <- makeBirthMatrix(flatBirthRates, tstep, lambda, a, w, m, v, b)
  B <- makeConstrainedBirthMatrix(flatBirthRates, tstep, lambda, a, w, m, v, b)
  res<-eigen(t(P+B))
  df_death_r[i] <- log(Re(res$values[1]))
  x_pi <- res$vectors[,1] / sum(res$vectors[,1])
  df_death[i,] <- Re(x_pi)
}
header <- paste('death', deaths, sep=' ')
rownames(df_death) <- header
rownames(df_death) <- header
df_death <- tbl_df(t(df_death))

df_death <- df_death %>% mutate("age"=seq(1,lambda))
df_death %>% gather(death, freq, -age) %>% ggplot() + aes(x=age,y=freq,color=death) + geom_line()

qplot(deaths, df_death_r, geom='line')
```


### High old mortality, flat birth rate

Plot different values of lambda (longevity) holding other things constant.

```{r}
lambdas <- seq(w, 5*w, w)
df_lambda <- matrix(0, nrow=length(lambdas), ncol=5*w)
df_lambda_r <- matrix(0, nrow=length(lambdas), ncol=1)
for (i in 1:length(lambdas)) {
  P <- makeSurvivalMatrix(makeDeathRates, tstep, lambdas[i], a, w, d, 5, .2)
  B <- makeConstrainedBirthMatrix(flatBirthRates, tstep, lambdas[i], a, w, m, v, b)
  res<-eigen(t(P+B))
  df_lambda_r[i] <- log(Re(res$values[1]))
  x_pi <- res$vectors[,1] / sum(res$vectors[,1])
  # df_lambda[i,] <- c(Re(x_pi), rep(0, 5*w-lambdas[i]))
  df_lambda[i,] <- c(Re(x_pi), rep(NA, 5*w-lambdas[i]))
}
header <- paste('lambda', lambdas, sep=' ')
rownames(df_lambda) <- header
df_lambda <- tbl_df(t(df_lambda))
df_lambda <- df_lambda %>% mutate("age"=seq(1, 5*w))
df_lambda %>% gather(lambda, freq, -age) %>% ggplot() + aes(x=age,y=freq,color=lambda) + geom_line()

qplot(lambdas, df_lambda_r, geom='line')
```

Plot different values of alpha (first reproduction age) holding other things
constant.

```{r}
alphastep <- 1
alphas <- seq(1, (w-1), alphastep)
df_alpha <- matrix(0, nrow=length(alphas), ncol=lambda)
df_alpha_r <- matrix(0, nrow=length(alphas), ncol=1)
for (i in 1:length(alphas)) {
  # P <- makeSurvivalMatrix(makeDeathRates, tstep, lambda, alphas[i], w, d, 5, .2)
  P <- makeSurvivalMatrix(makeDeathRates2, tstep, lambda, alphas[i], w, d, 5, .2)
  # B <- makeConstrainedBirthMatrix(flatBirthRates, tstep, lambda, alphas[i], w, m, v, b)
  B <- makeBirthMatrix(flatBirthRates, tstep, lambda, alphas[i], w, m, v, b)
  res<-eigen(t(P+B))
  df_alpha_r[i] <- log(Re(res$values[1]))
  x_pi <- res$vectors[,1] / sum(res$vectors[,1])
  df_alpha[i,] <- Re(x_pi)
}
header <- paste('alpha=0', alphas[1:ceiling(9/alphastep)], sep='')
header <- c(header, paste('alpha=', alphas[-(1:ceiling(9/alphastep))], sep=''))

rownames(df_alpha) <- header
colnames(df_alpha_r) <- 'r'
df_alpha <- tbl_df(t(df_alpha))

df_alpha <- df_alpha %>% mutate("age"=seq(1,lambda))
df_alpha %>% gather(alpha, freq, -age) %>% ggplot() + aes(x=age,y=freq,color=alpha) + geom_line()

df_alpha_r <- tbl_df(df_alpha_r) %>% mutate(alpha=alphas)
df_alpha_r %>% ggplot() + aes(x=alpha, y=r) + geom_line() + geom_hline(aes(yintercept=0))
```

Plot different 'menopause' distribution (omega) holding other things constant.

```{r}
omegastep <- 2
omegas <- seq(a+1, lambda, omegastep)
df_omega <- matrix(0, nrow=length(omegas), ncol=lambda)
df_omega_r <- matrix(0, nrow=length(omegas), ncol=1)
for (i in 1:length(omegas)) {
  # P <- makeSurvivalMatrix(makeDeathRates, tstep, lambda, a, omegas[1], d, 5, .2)
  P <- makeSurvivalMatrix(makeDeathRates2, tstep, lambda, a, omegas[i], d, 5, .2)
  # B <- makeConstrainedBirthMatrix(flatBirthRates, tstep, lambda, a, omegas[i], m, v, b)
  P <- makeSurvivalMatrix(makeDeathRates, tstep, lambda, a, omegas[i], d, 5, .2)
  res<-eigen(t(P+B))
  df_omega_r[i] <- log(Re(res$values[1]))
  x_pi <- res$vectors[,1] / sum(res$vectors[,1])
  df_omega[i,] <- Re(x_pi)
}
header <- paste('omega=', omegas, sep='')
rownames(df_omega) <- header
df_omega <- tbl_df(t(df_omega))

df_omega <- df_omega %>% mutate("age"=seq(1,lambda))
df_omega %>% gather(omega, freq, -age) %>% ggplot() + aes(x=age,y=freq,color=omega) + geom_line()

qplot(omegas, df_omega_r, geom='line')
```

Plot different death distribution holding other things constant.

```{r}
deaths <- seq(0, .45, .025)
df_death <- matrix(0, nrow=length(deaths), ncol=lambda)
df_death_r <- matrix(0, nrow=length(deaths), ncol=1)
for (i in 1:length(deaths)) {
  # P <- makeSurvivalMatrix(makeDeathRates, tstep, lambda, a, w, deaths[i], 5, .2)
  P <- makeSurvivalMatrix(makeDeathRates2, tstep, lambda, a, w, deaths[i], 5, .2)
  # B <- makeConstrainedBirthMatrix(flatBirthRates, tstep, lambda, a, w, m, v, b)
  B <- makeBirthMatrix(flatBirthRates, tstep, lambda, a, w, m, v, b)
  res<-eigen(t(P+B))
  df_death_r[i] <- log(Re(res$values[1]))
  x_pi <- res$vectors[,1] / sum(res$vectors[,1])
  df_death[i,] <- Re(x_pi)
}
header <- paste('death', deaths, sep=' ')
rownames(df_death) <- header
rownames(df_death) <- header
df_death <- tbl_df(t(df_death))

df_death <- df_death %>% mutate("age"=seq(1,lambda))
df_death %>% gather(death, freq, -age) %>% ggplot() + aes(x=age,y=freq,color=death) + geom_line()

qplot(deaths, df_death_r, geom='line')
```


# MCMC

Given a death distribution, find the distribution of alpha, omega, and lambda
parameters that maximize r.


## Uniform survival

```{r}
computeGrowthRate <- function(birthFxn, deathFxn, tstep, B, lambda, alpha,
                              omega, m, v, death, death_a, death_b) {
  # P <- makeSurvivalMatrix(deathFxn, tstep, lambda, d=death, death_a, death_b)
  # B <- makeBirthMatrix(birthFxn, tstep, lambda, alpha, omega, m, v, B)
  P <- makeSurvivalMatrix(deathFxn, tstep, lambda, alpha, omega, d=death, death_a, death_b)
  B <- makeConstrainedBirthMatrix(birthFxn, tstep, lambda, alpha, omega, m, v, B)
  res<-eigen(t(P+B))
  return(log(Re(res$values[1])))
}

updateParameter <- function(cur, move_m, move_sd, lower, upper, type) {
  if (type == 'int' ) {
    param <- cur + round(rnorm(1, mean=move_m, sd=move_sd)) # keep integer
  } else if (type == 'double') {
    param <- cur + rnorm(1, mean=move_m, sd=move_sd) # keep double
  }

  if (param > upper) {
    param <- updateParameter(cur, move_m, move_sd, lower, upper, type)
  } else if (param < lower) {
    param <- updateParameter(cur, move_m, move_sd, lower, upper, type)
  }
  # } else if (param == cur) {
  #   param <- updateParameter(cur, move_m, move_sd, lower, upper, type)
  # }

  return(param)
}

growthRateMCMC <- function(birthFxn, deathFxn, niter=1000, tstep=1, B0=20,
                           lambda0=50, alpha0=15, omega0=30, m0=20, v0=5,
                           death=.01, death_a=1, death_b=1) {
  d <- death
  d_a <- death_a
  d_b <- death_b

  lambdas <- rep(0, niter+1)
  alphas <- rep(0, niter+1)
  omegas <- rep(0, niter+1)
  ms <- rep(0, niter+1)
  vs <- rep(0, niter+1)
  Bs <- rep(0, niter+1)
  rs <- rep(0, niter+1)

  lambdas[1] <- lambda0
  alphas[1] <- alpha0
  omegas[1] <- omega0
  ms[1] <- m0
  vs[1] <- v0
  Bs[1] <- B0
  rs[1] <- computeGrowthRate(birthFxn, deathFxn, tstep, Bs[1], lambdas[1],
                             alphas[1], omegas[1], ms[1], vs[1], d, d_a, d_b)

  for (i in 2:(niter+1)) {
    current_r <- rs[i-1]

    # Update alphas
    # print('update alpha')
    proposed_a <- updateParameter(alphas[i-1], 0, 1, 1, omegas[i-1]-1, 'int')
    new_r <- computeGrowthRate(birthFxn, deathFxn, tstep, Bs[i-1],
                               lambdas[i-1], proposed_a, omegas[i-1], ms[i-1],
                               vs[i-1], d, d_a, d_b)
    A <- new_r/current_r
    if (runif(1) < A) {
      alphas[i] <- proposed_a
    } else {
      alphas[i] <- alphas[i-1]
    }
    
    # Update omegas
    # print('update omega')
    proposed_w <- updateParameter(omegas[i-1], 0, 1, alphas[i]+1, lambdas[i-1]-1, 'int')
    new_r <- computeGrowthRate(birthFxn, deathFxn, tstep, Bs[i-1],
                               lambdas[i-1], alphas[i], proposed_w, ms[i-1],
                               vs[i-1], d, d_a, d_b)
    #A <- new_r/current_r
    A <- (new_r-current_r)/current_r
    if (runif(1) < A) {
      omegas[i] <- proposed_w
    } else {
      omegas[i] <- omegas[i-1]
    }
    
    # Update lambdas
    # print('update lambda')
    proposed_l <- updateParameter(lambdas[i-1], 0, 1, omegas[i], 10*lambda0, 'int')
    new_r <- computeGrowthRate(birthFxn, deathFxn, tstep, Bs[i-1],
                               proposed_l, alphas[i], omegas[i], ms[i-1],
                               vs[i-1], d, d_a, d_b)
    A <- new_r/current_r
    if (runif(1) < A) {
      lambdas[i] <- proposed_l
    } else {
      lambdas[i] <- lambdas[i-1]
    }
    
    # Update means
    proposed_m <- updateParameter(ms[i-1], 0, .5, 0, lambdas[i], 'double')
    new_r <- computeGrowthRate(birthFxn, deathFxn, tstep, Bs[i-1],
                               lambdas[i], alphas[i], omegas[i], proposed_m,
                               vs[i-1], d, d_a, d_b)
    A <- new_r/current_r
    if (runif(1) < A) {
      ms[i] <- proposed_m
    } else {
      ms[i] <- ms[i-1]
    }
    
    # Update variances
    proposed_v <- updateParameter(vs[i-1], 0, .1, 0, lambdas[i], 'double')
    new_r <- computeGrowthRate(birthFxn, deathFxn, tstep, Bs[i-1],
                               lambdas[i], alphas[i], omegas[i], ms[i],
                               proposed_v, d, d_a, d_b)
    A <- new_r/current_r
    if (runif(1) < A) {
      vs[i] <- proposed_v
    } else {
      vs[i] <- vs[i-1]
    }
    
    # Update litter size
    proposed_B <- updateParameter(Bs[i-1], 0, 1, 1, 1000, 'int')
    new_r <- computeGrowthRate(birthFxn, deathFxn, tstep, proposed_B,
                               lambdas[i], alphas[i], omegas[i], ms[i], vs[i],
                               d, d_a, d_b)
    A <- new_r/current_r
    if (runif(1) < A) {
      Bs[i] <- proposed_B
    } else {
      Bs[i] <- Bs[i-1]
    }

    rs[i] <- computeGrowthRate(birthFxn, deathFxn, tstep, Bs[i], lambdas[i],
                               alphas[i], omegas[i], ms[i], vs[i], d, d_a,
                               d_b)
    print(rs[i])
    print(lambdas[i])
    print(alphas[i])
    print(omegas[i])
    print(Bs[i])
  }

  # return(list(r=rs, l=lambdas, a=alphas, w=omegas, m=ms, v=vs, B=Bs))
  return(list(r=rs, l=lambdas, a=alphas, w=omegas, B=Bs))
}
```

```{r}
niter <- 1000
tstep <- 1
B0 <- 10
lambda0 <- 50
alpha0 <- 15
omega0 <- 30
m0 <- 20
v0 <- 5
death <- .1
```

```{r}
# death_a <- 1
# death_b <- 1
# 
# res <- growthRateMCMC(flatBirthRates, makeDeathRates, niter, tstep, B0,
#                       lambda0, alpha0, omega0, m0, v0, death, death_a, death_b)
# df_res <- tbl_df(res)
```

```{r}
# df_res[which.max(res$r),]
```

```{r}
# ggplot(df_res) + aes(x=a, y=r) + geom_point()
# ggplot(df_res) + aes(x=B, y=r) + geom_point()
# ggplot(df_res) + aes(x=w, y=r) + geom_point()
# ggplot(df_res) + aes(x=l, y=r) + geom_point()
```


### Early mortality

```{r}
# death_a <- .2
# death_b <- 5
# 
# res <- growthRateMCMC(flatBirthRates, makeDeathRates, niter, tstep, B0,
#                       lambda0, alpha0, omega0, m0, v0, death, death_a, death_b)
# df_res <- tbl_df(res)
```

```{r}
# df_res[which.max(res$r),]
```


### Late mortality

```{r}
# death_a <- 5
# death_b <- .2
# 
# res <- growthRateMCMC(flatBirthRates, makeDeathRates, niter, tstep, B0,
#                       lambda0, alpha0, omega0, m0, v0, death, death_a, death_b)
# df_res <- tbl_df(res)
```

```{r}
# df_res[which.max(res$r),]
```
