---
title: Project
author: Ansel George
encoding: "UTF-8"
output: pdf_document
---

```{r}
library(expm)
library(ggplot2)
library(dplyr)
library(tidyr)

options(encoding = 'UTF-8')
set.seed(10)
```


# Functions

```{r}
# flatDeathRates <- function(t, d, a=1, b=1) {
#   return((1-d) * rep(1,length(t)))
# }

makeDeathRates <- function(t, d, a=1, b=1) {
  lambda <- max(t)+1
  p <- pbeta((t+1)/lambda, a, b) - pbeta(t/lambda, a, b)
  p <- 1-p
  p <- p/mean(p)
  return((1-d)*p)
  # p <- (1-d*dbeta((t+.1)/(max(t)+.1), a, b))
  # p <- dbeta((t+1)/(max(t)+1), a, b)
  # p <- (1-d*p/max(p))
}

makeSurvivalMatrix <- function(fxn, tstep, lambda, d=.1, a=1, b=1) {
  return(rbind(cbind(rep(0, lambda-1),
                     diag(lambda-1)),
               rep(0, lambda)) * fxn(seq(0, (lambda-1)*tstep), d, a, b))
}

# makeSurvivalMatrix <- function(fxn, tstep, lambda) {
#   d <- fxn(seq(0, (lambda)*tstep, tstep))
#   Q <- matrix(0, nrow=(lambda+1), ncol=(lambda+1))
#   Q <- -diag(lambda+1) + rbind(cbind(rep(0, lambda), diag(lambda)), rep(0, lambda+1)) %*% diag(1-d)
#   Q[,(lambda+1)] <- d
#   Q[(lambda+1),(lambda+1)] <- 0
#   Q[(lambda),(lambda+1)] <- 1
# 
#   return(expm(Q)[-(lambda+1),-(lambda+1)])
# }

flatBirthRates <- function(t, a, w, m, v) {
  b <- rep(0, length(t))
  b[a:w] <- 1
  return(b)
}

normalBirthRates <- function(age, a, w, m, v) {
  p <- dnorm(age, m, sqrt(v))
  return((w-a+1)*p/sum(p))
}

makeBirthMatrix <- function(fxn, tstep, lambda, a=0, w=0, m=1, v=0, b=1) {
  B <- matrix(0, nrow=lambda, ncol=lambda)
  B[1,] <- b*fxn(seq(0, (lambda-1)*tstep, tstep), a, w, m, v)
  return(t(B))
}

simulateGenerations <- function(x, P, B, n) {
  x_new <- x %*% ((P+B) %^% n)
  x_new <- x_new / sum(x_new)
  return(x_new)
  
  # X <- matrix(0, nrow=n, ncol=length(x))
  # N <- matrix(0, nrow=n, ncol=1)
  # X[1,] <- x
  # N[1] <- sum(x)
  # 
  # for (i in 2:n) {
  #   tmp <- X[i-1,] %*% (P+B)
  #   X[i,] <- tmp / sum(tmp)
  #   N[i] <- sum(tmp)
  # }
  # return(X)
}
```


# Global variables

```{r}
tstep <- 1
a <- 15
w <- 30
b <- 20
d <- .1
m <- 20
v <- 5
lambda <- 50

x <- rep(1/lambda, lambda)
```


# Simple simulations

## Flat birth/death rates

```{r}
P <- makeSurvivalMatrix(makeDeathRates, tstep, lambda, d, 1, 1)
B <- makeBirthMatrix(flatBirthRates, tstep, lambda, a, w, m, v, b)

res<-eigen(t(P+B))
log(res$values[1])
x_pi <- res$vectors[,1] / sum(res$vectors[,1])

age <- seq(1, lambda, tstep)
qplot(age, x)
qplot(age, Re(x_pi), geom='line')

log(x_pi[2])-log(x_pi[1])
log(x_pi[7])-log(x_pi[6])
```

Plot different values of lambda (longevity) holding other things constant.

```{r}
lambdas <- seq(w, 5*w, w)
df_lambda <- matrix(0, nrow=length(lambdas), ncol=5*w)
df_lambda_r <- matrix(0, nrow=length(lambdas), ncol=1)
for (i in 1:length(lambdas)) {
  P <- makeSurvivalMatrix(makeDeathRates, tstep, lambdas[i], d, 1, 1) 
  B <- makeBirthMatrix(flatBirthRates, tstep, lambdas[i], a, w, m, v, b)
  res<-eigen(t(P+B))
  df_lambda_r[i] <- log(Re(res$values[1]))
  x_pi <- res$vectors[,1] / sum(res$vectors[,1])
  # df_lambda[i,] <- c(Re(x_pi), rep(0, 5*w-lambdas[i]))
  df_lambda[i,] <- c(Re(x_pi), rep(NA, 5*w-lambdas[i]))
}
header <- paste('lambda', lambdas, sep=' ')
rownames(df_lambda) <- header
df_lambda <- tbl_df(t(df_lambda))
df_lambda <- df_lambda %>% mutate("age"=seq(1, 5*w))
df_lambda %>% gather(lambda, freq, -age) %>% ggplot() + aes(x=age,y=freq,color=lambda) + geom_line()

qplot(lambdas, df_lambda_r, geom='line')
```

Plot different values of alpha (first reproduction age) holding other things
constant.

```{r}
alphastep <- 2
alphas <- seq(1, (w-1), alphastep)
df_alpha <- matrix(0, nrow=length(alphas), ncol=lambda)
df_alpha_r <- matrix(0, nrow=length(alphas), ncol=1)
for (i in 1:length(alphas)) {
  P <- makeSurvivalMatrix(makeDeathRates, tstep, lambda, d, 1, 1)
  B <- makeBirthMatrix(flatBirthRates, tstep, lambda, alphas[i], w, m, v, b)
  res<-eigen(t(P+B))
  df_alpha_r[i] <- log(Re(res$values[1]))
  x_pi <- res$vectors[,1] / sum(res$vectors[,1])
  df_alpha[i,] <- Re(x_pi)
}
header <- paste('alpha=0', alphas[1:ceiling(9/alphastep)], sep='')
header <- c(header, paste('alpha=', alphas[-(1:ceiling(9/alphastep))], sep=''))

rownames(df_alpha) <- header
colnames(df_alpha_r) <- 'r'
df_alpha <- tbl_df(t(df_alpha))

df_alpha <- df_alpha %>% mutate("age"=seq(1,lambda))
df_alpha %>% gather(alpha, freq, -age) %>% ggplot() + aes(x=age,y=freq,color=alpha) + geom_line()

df_alpha_r <- tbl_df(df_alpha_r) %>% mutate(alpha=alphas)
df_alpha_r %>% ggplot() + aes(x=alpha, y=r) + geom_line() + geom_hline(aes(yintercept=0))
```

Plot different 'menopause' distribution (omega) holding other things constant.

```{r}
omegastep <- 2
omegas <- seq(a+1, lambda, omegastep)
df_omega <- matrix(0, nrow=length(omegas), ncol=lambda)
df_omega_r <- matrix(0, nrow=length(omegas), ncol=1)
for (i in 1:length(omegas)) {
  P <- makeSurvivalMatrix(makeDeathRates, tstep, lambda, d, 1, 1)
  B <- makeBirthMatrix(flatBirthRates, tstep, lambda, a, omegas[i], m, v, b)
  res<-eigen(t(P+B))
  df_omega_r[i] <- log(Re(res$values[1]))
  x_pi <- res$vectors[,1] / sum(res$vectors[,1])
  df_omega[i,] <- Re(x_pi)
}
header <- paste('omega=', omegas, sep='')
rownames(df_omega) <- header
df_omega <- tbl_df(t(df_omega))

df_omega <- df_omega %>% mutate("age"=seq(1,lambda))
df_omega %>% gather(omega, freq, -age) %>% ggplot() + aes(x=age,y=freq,color=omega) + geom_line()

qplot(omegas, df_omega_r, geom='line')
```

Plot different death distribution holding other things constant.

```{r}
deaths <- seq(0, .95, .05)
df_death <- matrix(0, nrow=length(deaths), ncol=lambda)
df_death_r <- matrix(0, nrow=length(deaths), ncol=1)
for (i in 1:length(deaths)) {
  P <- makeSurvivalMatrix(makeDeathRates, tstep, lambda, deaths[i], 1, 1)
  B <- makeBirthMatrix(flatBirthRates, tstep, lambda, a, w, m, v, b)
  res<-eigen(t(P+B))
  df_death_r[i] <- log(Re(res$values[1]))
  x_pi <- res$vectors[,1] / sum(res$vectors[,1])
  df_death[i,] <- Re(x_pi)
}
header <- paste('death', deaths, sep=' ')
rownames(df_death) <- header
rownames(df_death) <- header
df_death <- tbl_df(t(df_death))

df_death <- df_death %>% mutate("age"=seq(1,lambda))
df_death %>% gather(death, freq, -age) %>% ggplot() + aes(x=age,y=freq,color=death) + geom_line()

qplot(deaths, df_death_r, geom='line')
```


## Normal birth, flat death

Plot different values of lambda (longevity) holding other things constant.

```{r}
lambdas <- seq(m, 3*m, 2)
t_lambda <- matrix(0, nrow=length(lambdas), ncol=3*m)
t_lambda_r <- matrix(0, nrow=length(lambdas), ncol=1)
for (i in 1:length(lambdas)) {
  P <- makeSurvivalMatrix(makeDeathRates, tstep, lambdas[i], d, 1, 1)
  B <- makeBirthMatrix(normalBirthRates, tstep, lambdas[i], a, w, m, v, b)
  res<-eigen(t(P+B))
  t_lambda_r[i] <- log(Re(res$values[1]))
  x_pi <- res$vectors[,1] / sum(res$vectors[,1])
  t_lambda[i,] <- c(Re(x_pi), rep(NA, 3*m-lambdas[i]))
}
header <- paste('lambda', lambdas, sep=' ')
rownames(t_lambda) <- header
df_lambda <- tbl_df(t(t_lambda))
df_lambda <- df_lambda %>% mutate("age"=seq(1, 3*m))
df_lambda %>% gather(lambda, freq, -age) %>% ggplot() + aes(x=age,y=freq,color=lambda) + geom_line()

qplot(lambdas, t_lambda_r, geom='line')
qplot(x=seq(1, length(B[,1])), B[,1])
```

Plot different values of mean reproduction age (m) holding other things
constant.

```{r}
meanstep <- 1
means <- seq(ceiling(m/2), 2*m, meanstep)
df_mean <- matrix(0, nrow=length(means), ncol=lambda)
df_mean_r <- matrix(0, nrow=length(means), ncol=1)
for (i in 1:length(means)) {
  P <- makeSurvivalMatrix(makeDeathRates, tstep, lambda, d, 1, 1)
  B <- makeBirthMatrix(normalBirthRates, tstep, lambda, a, w, means[i], v, b)
  res<-eigen(t(P+B))
  df_mean_r[i] <- log(Re(res$values[1]))
  x_pi <- res$vectors[,1] / sum(res$vectors[,1])
  df_mean[i,] <- Re(x_pi)
}
header <- paste('m=', means, sep='')

rownames(df_mean) <- header
colnames(df_mean_r) <- 'r'
df_mean <- tbl_df(t(df_mean))

df_mean <- df_mean %>% mutate("age"=seq(1,lambda))
df_mean %>% gather(m, freq, -age) %>% ggplot() + aes(x=age,y=freq,color=m) + geom_line()

df_mean_r <- tbl_df(df_mean_r) %>% mutate(m=means)
df_mean_r %>% ggplot() + aes(x=m, y=r) + geom_line() + geom_hline(aes(yintercept=0), color='red')
```

Plot variance of different distributions other things constant.

```{r}
vstep <- 1
vs <- seq(2, 10)
t_v <- matrix(0, nrow=length(vs), ncol=lambda)
t_v_r <- matrix(0, nrow=length(vs), ncol=1)
for (i in 1:length(vs)) {
  P <- makeSurvivalMatrix(makeDeathRates, tstep, lambda, d, 1, 1)
  B <- makeBirthMatrix(normalBirthRates, tstep, lambda, a, w, m, vs[i], b)
  res <- eigen(t(P+B))
  t_v_r[i] <- log(Re(res$values[1]))
  x_pi <- res$vectors[,1] / sum(res$vectors[,1])
  t_v[i,] <- Re(x_pi)
}
header <- paste('v=0', vs[1:8], sep='')
header <- c(header, paste('v=', vs[-(1:8)], sep=''))

rownames(t_v) <- header
colnames(t_v_r) <- 'r'
df_v <- tbl_df(t(t_v))

df_v <- df_v %>% mutate("age"=seq(1,lambda))
df_v %>% gather(v, freq, -age) %>% ggplot() + aes(x=age,y=freq,color=v) + geom_line()

df_v_r <- tbl_df(t_v_r) %>% mutate(v=vs)
df_v_r %>% ggplot() + aes(x=v, y=r) + geom_line() + geom_hline(aes(yintercept=0))
```

Plot different death distribution holding other things constant.

```{r}
deaths <- seq(0, .95, .05)
df_death <- matrix(0, nrow=length(deaths), ncol=lambda)
df_death_r <- matrix(0, nrow=length(deaths), ncol=1)
for (i in 1:length(deaths)) {
  P <- makeSurvivalMatrix(makeDeathRates, tstep, lambda, deaths[i], 1, 1)
  B <- makeBirthMatrix(normalBirthRates, tstep, lambda, a, w, m, v, b)
  res<-eigen(t(P+B))
  df_death_r[i] <- log(Re(res$values[1]))
  x_pi <- res$vectors[,1] / sum(res$vectors[,1])
  df_death[i,] <- Re(x_pi)
}
header <- paste('death=', deaths, sep='')
rownames(df_death) <- header
rownames(df_death) <- header
df_death <- tbl_df(t(df_death))

df_death <- df_death %>% mutate("age"=seq(1,lambda))
df_death %>% gather(death, freq, -age) %>% ggplot() + aes(x=age,y=freq,color=death) + geom_line()

qplot(deaths, df_death_r, geom='line')
```

## Add skewed survivorship function

### High young mortality, flat birth rate 

Plot different values of lambda (longevity) holding other things constant.

```{r}
lambdas <- seq(w, 5*w, w)
df_lambda <- matrix(0, nrow=length(lambdas), ncol=5*w)
df_lambda_r <- matrix(0, nrow=length(lambdas), ncol=1)
for (i in 1:length(lambdas)) {
  P <- makeSurvivalMatrix(makeDeathRates, tstep, lambdas[i], d, .5, 2)
  B <- makeBirthMatrix(flatBirthRates, tstep, lambdas[i], a, w, m, v, b)
  res<-eigen(t(P+B))
  df_lambda_r[i] <- log(Re(res$values[1]))
  x_pi <- res$vectors[,1] / sum(res$vectors[,1])
  # df_lambda[i,] <- c(Re(x_pi), rep(0, 5*w-lambdas[i]))
  df_lambda[i,] <- c(Re(x_pi), rep(NA, 5*w-lambdas[i]))
}
header <- paste('lambda', lambdas, sep=' ')
rownames(df_lambda) <- header
df_lambda <- tbl_df(t(df_lambda))
df_lambda <- df_lambda %>% mutate("age"=seq(1, 5*w))
df_lambda %>% gather(lambda, freq, -age) %>% ggplot() + aes(x=age,y=freq,color=lambda) + geom_line()

qplot(lambdas, df_lambda_r, geom='line')
```

Plot different values of alpha (first reproduction age) holding other things
constant.

```{r}
alphastep <- 2
alphas <- seq(1, (w-1), alphastep)
df_alpha <- matrix(0, nrow=length(alphas), ncol=lambda)
df_alpha_r <- matrix(0, nrow=length(alphas), ncol=1)
for (i in 1:length(alphas)) {
  P <- makeSurvivalMatrix(makeDeathRates, tstep, lambda, d, .5, 2)
  B <- makeBirthMatrix(flatBirthRates, tstep, lambda, alphas[i], w, m, v, b)
  res<-eigen(t(P+B))
  df_alpha_r[i] <- log(Re(res$values[1]))
  x_pi <- res$vectors[,1] / sum(res$vectors[,1])
  df_alpha[i,] <- Re(x_pi)
}
header <- paste('alpha=0', alphas[1:ceiling(9/alphastep)], sep='')
header <- c(header, paste('alpha=', alphas[-(1:ceiling(9/alphastep))], sep=''))

rownames(df_alpha) <- header
colnames(df_alpha_r) <- 'r'
df_alpha <- tbl_df(t(df_alpha))

df_alpha <- df_alpha %>% mutate("age"=seq(1,lambda))
df_alpha %>% gather(alpha, freq, -age) %>% ggplot() + aes(x=age,y=freq,color=alpha) + geom_line()

df_alpha_r <- tbl_df(df_alpha_r) %>% mutate(alpha=alphas)
df_alpha_r %>% ggplot() + aes(x=alpha, y=r) + geom_line() + geom_hline(aes(yintercept=0))
```

Plot different 'menopause' distribution (omega) holding other things constant.

```{r}
omegastep <- 2
omegas <- seq(a+1, lambda, omegastep)
df_omega <- matrix(0, nrow=length(omegas), ncol=lambda)
df_omega_r <- matrix(0, nrow=length(omegas), ncol=1)
for (i in 1:length(omegas)) {
  P <- makeSurvivalMatrix(makeDeathRates, tstep, lambda, d, .5, 2)
  # P <- makeSurvivalMatrix(makeDeathRates, tstep, lambda, d, 1, 1)
  B <- makeBirthMatrix(flatBirthRates, tstep, lambda, a, omegas[i], m, v, b)
  res<-eigen(t(P+B))
  df_omega_r[i] <- log(Re(res$values[1]))
  x_pi <- res$vectors[,1] / sum(res$vectors[,1])
  df_omega[i,] <- Re(x_pi)
}
header <- paste('omega=', omegas, sep='')
rownames(df_omega) <- header
df_omega <- tbl_df(t(df_omega))

df_omega <- df_omega %>% mutate("age"=seq(1,lambda))
df_omega %>% gather(omega, freq, -age) %>% ggplot() + aes(x=age,y=freq,color=omega) + geom_line()

qplot(omegas, df_omega_r, geom='line')
```

Plot different death distribution holding other things constant.

```{r}
deaths <- seq(0, .95, .05)
df_death <- matrix(0, nrow=length(deaths), ncol=lambda)
df_death_r <- matrix(0, nrow=length(deaths), ncol=1)
for (i in 1:length(deaths)) {
  P <- makeSurvivalMatrix(makeDeathRates, tstep, lambda, deaths[i], .5, 2)
  B <- makeBirthMatrix(flatBirthRates, tstep, lambda, a, w, m, v, b)
  res<-eigen(t(P+B))
  df_death_r[i] <- log(Re(res$values[1]))
  x_pi <- res$vectors[,1] / sum(res$vectors[,1])
  df_death[i,] <- Re(x_pi)
}
header <- paste('death', deaths, sep=' ')
rownames(df_death) <- header
rownames(df_death) <- header
df_death <- tbl_df(t(df_death))

df_death <- df_death %>% mutate("age"=seq(1,lambda))
df_death %>% gather(death, freq, -age) %>% ggplot() + aes(x=age,y=freq,color=death) + geom_line()

qplot(deaths, df_death_r, geom='line')
```


### High old mortality, flat birth rate

Plot different values of lambda (longevity) holding other things constant.

```{r}
lambdas <- seq(w, 5*w, w)
df_lambda <- matrix(0, nrow=length(lambdas), ncol=5*w)
df_lambda_r <- matrix(0, nrow=length(lambdas), ncol=1)
for (i in 1:length(lambdas)) {
  P <- makeSurvivalMatrix(makeDeathRates, tstep, lambdas[i], d, 2, .5)
  B <- makeBirthMatrix(flatBirthRates, tstep, lambdas[i], a, w, m, v, b)
  res<-eigen(t(P+B))
  df_lambda_r[i] <- log(Re(res$values[1]))
  x_pi <- res$vectors[,1] / sum(res$vectors[,1])
  # df_lambda[i,] <- c(Re(x_pi), rep(0, 5*w-lambdas[i]))
  df_lambda[i,] <- c(Re(x_pi), rep(NA, 5*w-lambdas[i]))
}
header <- paste('lambda', lambdas, sep=' ')
rownames(df_lambda) <- header
df_lambda <- tbl_df(t(df_lambda))
df_lambda <- df_lambda %>% mutate("age"=seq(1, 5*w))
df_lambda %>% gather(lambda, freq, -age) %>% ggplot() + aes(x=age,y=freq,color=lambda) + geom_line()

qplot(lambdas, df_lambda_r, geom='line')
```

Plot different values of alpha (first reproduction age) holding other things
constant.

```{r}
alphastep <- 2
alphas <- seq(1, (w-1), alphastep)
df_alpha <- matrix(0, nrow=length(alphas), ncol=lambda)
df_alpha_r <- matrix(0, nrow=length(alphas), ncol=1)
for (i in 1:length(alphas)) {
  P <- makeSurvivalMatrix(makeDeathRates, tstep, lambda, d, 2, .5)
  B <- makeBirthMatrix(flatBirthRates, tstep, lambda, alphas[i], w, m, v, b)
  res<-eigen(t(P+B))
  df_alpha_r[i] <- log(Re(res$values[1]))
  x_pi <- res$vectors[,1] / sum(res$vectors[,1])
  df_alpha[i,] <- Re(x_pi)
}
header <- paste('alpha=0', alphas[1:ceiling(9/alphastep)], sep='')
header <- c(header, paste('alpha=', alphas[-(1:ceiling(9/alphastep))], sep=''))

rownames(df_alpha) <- header
colnames(df_alpha_r) <- 'r'
df_alpha <- tbl_df(t(df_alpha))

df_alpha <- df_alpha %>% mutate("age"=seq(1,lambda))
df_alpha %>% gather(alpha, freq, -age) %>% ggplot() + aes(x=age,y=freq,color=alpha) + geom_line()

df_alpha_r <- tbl_df(df_alpha_r) %>% mutate(alpha=alphas)
df_alpha_r %>% ggplot() + aes(x=alpha, y=r) + geom_line() + geom_hline(aes(yintercept=0))
```

Plot different 'menopause' distribution (omega) holding other things constant.

```{r}
omegastep <- 2
omegas <- seq(a+1, lambda, omegastep)
df_omega <- matrix(0, nrow=length(omegas), ncol=lambda)
df_omega_r <- matrix(0, nrow=length(omegas), ncol=1)
for (i in 1:length(omegas)) {
  P <- makeSurvivalMatrix(makeDeathRates, tstep, lambda, d, 2, .5)
  B <- makeBirthMatrix(flatBirthRates, tstep, lambda, a, omegas[i], m, v, b)
  res<-eigen(t(P+B))
  df_omega_r[i] <- log(Re(res$values[1]))
  x_pi <- res$vectors[,1] / sum(res$vectors[,1])
  df_omega[i,] <- Re(x_pi)
}
header <- paste('omega=', omegas, sep='')
rownames(df_omega) <- header
df_omega <- tbl_df(t(df_omega))

df_omega <- df_omega %>% mutate("age"=seq(1,lambda))
df_omega %>% gather(omega, freq, -age) %>% ggplot() + aes(x=age,y=freq,color=omega) + geom_line()

qplot(omegas, df_omega_r, geom='line')
```

Plot different death distribution holding other things constant.

```{r}
deaths <- seq(0, .95, .05)
df_death <- matrix(0, nrow=length(deaths), ncol=lambda)
df_death_r <- matrix(0, nrow=length(deaths), ncol=1)
for (i in 1:length(deaths)) {
  P <- makeSurvivalMatrix(makeDeathRates, tstep, lambda, deaths[i], 2, .5)
  B <- makeBirthMatrix(flatBirthRates, tstep, lambda, a, w, m, v, b)
  res<-eigen(t(P+B))
  df_death_r[i] <- log(Re(res$values[1]))
  x_pi <- res$vectors[,1] / sum(res$vectors[,1])
  df_death[i,] <- Re(x_pi)
}
header <- paste('death', deaths, sep=' ')
rownames(df_death) <- header
rownames(df_death) <- header
df_death <- tbl_df(t(df_death))

df_death <- df_death %>% mutate("age"=seq(1,lambda))
df_death %>% gather(death, freq, -age) %>% ggplot() + aes(x=age,y=freq,color=death) + geom_line()

qplot(deaths, df_death_r, geom='line')
```


# MCMC

Given a death distribution, find the distribution of alpha, omega, and lambda
parameters that maximize r.


## Uniform survival

```{r}

computeGrowthRate <- function(birthFxn, deathFxn, tstep, B, lambda, alpha,
                              omega, m, v, death, death_a, death_b) {
  P <- makeSurvivalMatrix(deathFxn, tstep, lambda, d=death, death_a, death_b)
  B <- makeBirthMatrix(birthFxn, tstep, lambda, alpha, omega, m, v, B)
  res<-eigen(t(P+B))
  return(log(Re(res$values[1])))
}

updateParameter <- function(cur, move_m, move_sd, lower, upper, type) {
  if (type == 'int' ) {
    param <- cur + round(rnorm(1, mean=move_m, sd=move_sd)) # keep integer
  } else if (type == 'double') {
    param <- cur + rnorm(1, mean=move_m, sd=move_sd) # keep double
  }

  if (param > upper) {
    param <- updateParameter(cur, move_m, move_sd, lower, upper, type)
  } else if (param < lower) {
    param <- updateParameter(cur, move_m, move_sd, lower, upper, type)
  }
  # } else if (param == cur) {
  #   param <- updateParameter(cur, move_m, move_sd, lower, upper, type)
  # }

  return(param)
}

growthRateMCMC <- function(birthFxn, deathFxn, niter=1000, tstep=1, B0=20,
                           lambda0=50, alpha0=15, omega0=30, m0=20, v0=5,
                           death=.01, death_a=1, death_b=1) {
  d <- death
  d_a <- death_a
  d_b <- death_b

  lambdas <- rep(0, niter+1)
  alphas <- rep(0, niter+1)
  omegas <- rep(0, niter+1)
  ms <- rep(0, niter+1)
  vs <- rep(0, niter+1)
  Bs <- rep(0, niter+1)
  rs <- rep(0, niter+1)

  lambdas[1] <- lambda0
  alphas[1] <- alpha0
  omegas[1] <- omega0
  ms[1] <- m0
  vs[1] <- v0
  Bs[1] <- B0
  rs[1] <- computeGrowthRate(birthFxn, deathFxn, tstep, Bs[1], lambdas[1],
                             alphas[1], omegas[1], ms[1], vs[1], d, d_a, d_b)

  for (i in 2:(niter+1)) {
    current_r <- rs[i-1]

    # Update alphas
    # print('update alpha')
    proposed_a <- updateParameter(alphas[i-1], 0, 1, 1, omegas[i-1]-1, 'int')
    new_r <- computeGrowthRate(birthFxn, deathFxn, tstep, Bs[i-1],
                               lambdas[i-1], proposed_a, omegas[i-1], ms[i-1],
                               vs[i-1], d, d_a, d_b)
    A <- new_r/current_r
    if (runif(1) < A) {
      alphas[i] <- proposed_a
    } else {
      alphas[i] <- alphas[i-1]
    }
    
    # Update omegas
    # print('update omega')
    proposed_w <- updateParameter(omegas[i-1], 0, 1, alphas[i], lambdas[i-1]-1, 'int')
    new_r <- computeGrowthRate(birthFxn, deathFxn, tstep, Bs[i-1],
                               lambdas[i-1], alphas[i], proposed_w, ms[i-1],
                               vs[i-1], d, d_a, d_b)
    A <- new_r/current_r
    if (runif(1) < A) {
      omegas[i] <- proposed_w
    } else {
      omegas[i] <- omegas[i-1]
    }
    
    # Update lambdas
    # print('update lambda')
    proposed_l <- updateParameter(lambdas[i-1], 0, 1, omegas[i], 10*lambda0, 'int')
    new_r <- computeGrowthRate(birthFxn, deathFxn, tstep, Bs[i-1],
                               proposed_l, alphas[i], omegas[i], ms[i-1],
                               vs[i-1], d, d_a, d_b)
    A <- new_r/current_r
    if (runif(1) < A) {
      lambdas[i] <- proposed_l
    } else {
      lambdas[i] <- lambdas[i-1]
    }
    
    # Update means
    proposed_m <- updateParameter(ms[i-1], 0, .5, 0, lambdas[i], 'double')
    new_r <- computeGrowthRate(birthFxn, deathFxn, tstep, Bs[i-1],
                               lambdas[i], alphas[i], omegas[i], proposed_m,
                               vs[i-1], d, d_a, d_b)
    A <- new_r/current_r
    if (runif(1) < A) {
      ms[i] <- proposed_m
    } else {
      ms[i] <- ms[i-1]
    }
    
    # Update variances
    proposed_v <- updateParameter(vs[i-1], 0, .1, 0, lambdas[i], 'double')
    new_r <- computeGrowthRate(birthFxn, deathFxn, tstep, Bs[i-1],
                               lambdas[i], alphas[i], omegas[i], ms[i],
                               proposed_v, d, d_a, d_b)
    A <- new_r/current_r
    if (runif(1) < A) {
      vs[i] <- proposed_v
    } else {
      vs[i] <- vs[i-1]
    }
    
    # Update litter size
    proposed_B <- updateParameter(Bs[i-1], 0, 1, 0, 1000, 'int')
    # print(proposed_B)
    # print(Bs[i-1])

    new_r <- computeGrowthRate(birthFxn, deathFxn, tstep, proposed_B,
                               lambdas[i], alphas[i], omegas[i], ms[i], vs[i],
                               d, d_a, d_b)
    # print(new_r)
    # print(current_r)
    A <- new_r/current_r
    # print(A)
    if (runif(1) < A) {
      Bs[i] <- proposed_B
    } else {
      Bs[i] <- Bs[i-1]
    }

    rs[i] <- computeGrowthRate(birthFxn, deathFxn, tstep, Bs[i], lambdas[i],
                               alphas[i], proposed_w, ms[i], vs[i], d, d_a,
                               d_b)
  }

  return(list(r=rs, l=lambdas, a=alphas, w=omegas, m=ms, v=vs, B=Bs))
}
```

```{r}
niter <- 100
tstep <- 1
B0 <- 5
lambda0 <- 50
alpha0 <- 15
omega0 <- 30
m0 <- 20
v0 <- 5
death <- .01
death_a <- 1
death_b <- 1
```

```{r}
res <- growthRateMCMC(flatBirthRates, makeDeathRates, niter, tstep, B0,
                      lambda0, alpha0, omega0, m0, v0, death, death_a, death_b)
df_res <- tbl_df(res)
```

```{r}
df_res[which.max(res$r),]
```


### Early mortality

```{r}
death_a <- .5
death_b <- 2
```

```{r}
res <- growthRateMCMC(flatBirthRates, makeDeathRates, niter, tstep, B0,
                      lambda0, alpha0, omega0, m0, v0, death, death_a, death_b)
df_res <- tbl_df(res)
```

```{r}
df_res[which.max(res$r),]
```
