---
title: "Problem Set 6"
author: "Ansel George"
output: pdf_document
---

```{r}
library(gtools)

set.seed(10)
```

Suppose you have a number of samples from a 'population' that contains an
unknown fraction of forest or savannah elephants. (Assume elephants are haploid
for simplicity.) Introduce the following notation.

* Let $x_i$ denote the genetic data from elephant $i$. Each $x_i$ is a vector
  of $0$s and $1$s indicating which allele is observed at each of $J$ loci.
  (so $x_{ij} \in \{0,1\}$ is the allele observed in sample $i$ at locus $j$.)
* Let $z_i$ denote the indicator for whether elephant $i$ is from forest or
  savanna.
* Let $\pi_0$ denote the fraction of the population that is forest.
* Let $f_{0j}$ denote the frequency of allele "1" in forest elephants at locus
  $j$.
* Let $f_{1j}$ denote the frequency of allele "1" in savannah elephants at
  locus $j$.

Do the following:

# 1.
*Draw a graphical representation of the way these variables/parameters connect
with one another in a probability model. Identify which quantities are
observed/known and which are unobserved/unknown.*

```{r}
library(igraph)

adj <- rbind(c(0, 1, 1, 1, 0, 0, 0, 0),
             c(0, 0, 0, 0, 1, 0, 0, 0),
             c(0, 0, 0, 0, 0, 1, 0, 0),
             c(0, 0, 0, 0, 0, 0, 1, 0),
             c(0, 0, 0, 0, 0, 0, 0, 0),
             c(0, 0, 0, 0, 0, 0, 0, 0),
             c(0, 0, 0, 0, 0, 0, 0, 0),
             c(0, 0, 0, 0, 1, 1, 1, 0))

names <- c("Pi", "Z1", "Z2", "Z3", "X1", "X2", "X3", "P")
plot(graph_from_adjacency_matrix(adj, weighted=T), vertex.label=names)
```

Observed/Known variables:

* $P$ - allele frequencies for savannah and forest elephants, copied from
  \url{https://stephens999.github.io/fiveMinuteStats/likelihood_ratio_simple_models.html}
* $X$ - genotype data

Unknowns:

* $\Pi$ - overall frequency of forest vs savannah elephants
* $Z$ - origin of all elephants in the sample $X$


# 2.
*Write down the likelihood you would use to do inference for the unknowns.*

Based on the structure of the Bayesian network, frequencies of hidden states
$\Pi$ and allele frequencies per locus per population $P$ are independent of
other parameters. Also, data $X$ is conditionally independent of $\Pi$ given
hidden states $Z$.

\begin{align}
\implies P(X,\Pi,Z,P) &= P(\Pi)P(P)P(Z|\Pi)P(X|Z,P)
\end{align}


# 3.
*Take the unknown quantities one at a time, and temporarily imagine that you
were given the values for all the other variables. Write down the likelihood
for that unknown quantity, ignoring parts that are constants (i.e. do not
depend of that unknown). Using this, work out what the conjugate prior families
would be for each of the unknown quantities (if all other quantities were
known).*

Given full data for every parameter but one, the conditional distribution of
the one unknown can be computed by marginalizing over all other variables. In
Bayesian terms, $P(A|B) = \frac{P(A,B)}{P(B)} \propto P(A,B) \propto
\frac{P(B|A)}{P(A)}$. 


When $\Pi$ is unknown:

\begin{align}
P(\Pi|X,Z,P) &\propto P(\Pi)P(P)P(Z|\Pi)P(X|Z,P) \\
  &= P(\Pi)P(P)P(Z|\Pi)P(X|Z,P) \\
  &\propto P(\Pi)P(Z|\Pi)
\end{align}

The posterior distribution for $\Pi$ is proportional to the prior on $\Pi$ and
the likelihood of $Z|\Pi$.

$Z|\Pi = \prod_{i} \prod_{k} \pi_{k}^{z_i}$, where $z_i$ is an indicator $z \in
\{0,1\}$ for the population of origin for individual $i$, and $\pi_k$ being the
likelihood of an individual from subpopulation $k$. This is a multinomial
distribution and has a Dirichlet conjugate pair.

With a multinomial prior, both probabilities are multinomially-distributed (or
binomially in the case $k=2$), with a Dirichlet posterior distribution.
Initializing $\Pi$ with a uniform distribution is a simple instance of a
Dirichlet distribution.

\begin{align}
P(Z|X,\Pi,P) &\propto P(\Pi)P(P)P(Z|\Pi)P(X|Z,P) \\
  &\propto P(X|Z,P)P(Z|\Pi)
\end{align}

$P(X|Z,P)$ is the probability of observing genotype $x_i$ given known
individual origin $Z$ and subpopulation frequency $\Pi$. Note that each locus
is independent and therefore can be considered individually.

$P(x_i|Z,P) = \prod_j (\pi_{z_k})^{x_{ij}}(1-\pi_{z_k})^{1-x_{ij}}$, where the
product is over the $j$ loci in individual $x_i$ and $x_{ij}$ denotes the
genotype of individual $x_i$ at locus $j$. This likelihood, when normalized,
follows a Beta distribution.

The probability of $Z|\Pi$ follows a multinomial distribution. The Beta
distribution is a special case of a Dirichlet distribution, just as a
multinomial distribution is a generalization of the binomial. The posterior is
therefore Dirichlet-distributed.

\begin{align}
P(P|X,Z,\Pi) &\propto P(\Pi)P(P)P(Z|\Pi)P(X|Z,P) \\
  &\propto P(X|Z,P)P(P)
\end{align}

As explained above, the $P(x_i|Z,P)$ follows a binomial distribution with Beta
conjugate pair. When $\P$ is beta-distributed (assume uniform $\beta(1,1)$ for
simplicity), the posterior follows a Beta distribution.


# 4.
*Provide a detailed outline of a Gibbs sampler to fit the model. (This model is
analogous to the Structure model without admixture, where $k=2$.)*

The Gibbs sampler takes elephant genotype data $X$ and allele frequencies $P$
if known.

**Initialize:**

Initialize unknown parameters $\Pi^{(0)}$, $Z^{(0)}$, and $P^{(0)}$. 

$\Pi^{(0)} \sim Dir(\underline{\alpha}=1)$ (uniform), $Z^{(0)}$ is a random
sampling of $\Pi^{(0)}$, and $P^{(0)}$ is computed from the.

**Induction:**

Compute the conditional distributions based on the existing data and sample the
next iterations.

$\Pi^{(i+1)} = \Pi^{(i)} P(Z^{(i)}|\Pi^{(i)}) \propto \Pi^{(i)}
Dir(\underline{\alpha})$ where $\underline{\alpha}$ is a vector containing the
number of individuals in each subpopulation $k$. The new estimate for each
subpopulation is sampled from the Dirichlet posterior.

$P^{(i+1)} = P^{(i)} P(X|Z^{(i)}, P^{(i)})$

$Z^{(i+1)} = P(Z^{(0)}|\Pi^{(1)}) P(X|Z^{(0)}, P^{(1)})$

**Termination:**

Continue computing updates to $P$, $\Pi$, and $Z$ until sufficiently many
iterations have passed.


# 5.
*Implement this Gibbs sampler in either R or python.*

```{r}
#' @param x an R vector of data
#' @param P a K by R matrix of allele frequencies
#' @return the log-likelihood for each of the K populations
log_pr_x_given_P <- function(x,P){
  tP <- t(P) #transpose P so tP is R by K
  return(colSums(x*log(tP)+(1-x)*log(1-tP)))
}

normalize <- function(x){return(x/sum(x))} #used in sample_z below

#' @param x an n by R matrix of data
#' @param P a K by R matrix of allele frequencies
#' @return an n vector of group memberships
sample_z <- function(x, P, Pi){
  K <- nrow(P)
  loglik_matrix <- apply(x, 1, log_pr_x_given_P, P=P)

  lik_matrix <- exp(loglik_matrix)
  lik_matrix <- lik_matrix * as.vector(Pi)

  p.z.given.x <- apply(lik_matrix,2,normalize) # normalize columns

  z <- rep(0, nrow(x))
  for(i in 1:length(z)){
    z[i] <- sample(1:K, size=1,prob=p.z.given.x[,i],replace=TRUE)
  }
  return(z)
}

#' @param x an n by R matrix of data
#' @param z an n vector of cluster allocations
#' @return a k by R matrix of allele frequencies
sample_P <- function(x,z,k){
  R <- ncol(x)
  P <- matrix(ncol=R,nrow=k)
  for(i in 1:k){
    sample_size <- sum(z==i)
    if(sample_size==0){
      number_of_ones=rep(0,R) 
    } else {
      number_of_ones <- colSums(x[z==i,])
    }
    P[i,] <- rbeta(R,1+number_of_ones,1+sample_size-number_of_ones) 
  }
  return(P)
}

#' @param z an n vector of cluster allocations (1...k)
#' @param k the number of clusters
sample_pi <- function(z,k){
  counts <- colSums(outer(z,1:k,FUN="=="))
  Pi <- gtools::rdirichlet(1,counts+1)
  return(Pi)
}

gibbs <- function(x, P=NULL, k = 2, niter = 100){
  # Initialize priors
  z <- sample(1:k,nrow(x),replace=TRUE)
  Pi <- rep(1/k,k)

  res <- list(z = matrix(nrow=niter, ncol=nrow(x)), Pi = matrix(nrow=niter, ncol=k))
  res$z[1,]=z
  res$Pi[1,]=Pi
 
  pknown <- TRUE
  if (is.null(P)) {
    pknown <- FALSE
  }

  for(i in 2:niter){
    Pi <- sample_pi(z,k)
    if (pknown == FALSE) {
      P <- sample_P(x,z,k)
    }
    z <- sample_z(x,P,Pi)
    res$z[i,] <- z
    res$Pi[i,] <- Pi
  }
  if (pknown == FALSE) {
    print(P)
  }
  return(res)
}
```

# 6.
*Test your implementation on simulated data. Compare the results of your
algorithm with the true parameter values.*

```{r}
# Generate the genotype and indicator for an elephant.
makeElephant <- function(z, f) {
  k <- length(z) # number of subpopulations

  subpopulation <- sample(seq(1, k), 1, prob=z) 
  elephant <- rbinom(f[subpopulation,], 1, f[subpopulation,])
  return(list(x=elephant, z=subpopulation))
}

# Generate genotypes and indicators for n elephants.
makeElephants <- function(n, z, f) {
  genotypes <- matrix(0, n, ncol(f))
  indicators <- matrix(n, 0, 1)
  for (i in 1:n) {
    elephant <- makeElephant(z, f)
    genotypes[i,] <- elephant$x
    indicators[i] <- elephant$z
  }
  return(list(x=genotypes, z=indicators))
}
```

Using the provided allele frequencies for the savannah and forest elephants:

```{r}
n <- 1000
fs <- c(0.40, 0.12, 0.21, 0.12, 0.02, 0.32)
ff <- c(0.80, 0.20, 0.11, 0.17, 0.23, 0.25)
f <- rbind(fs, ff)
p <- c(.2, .8) # 20% savannah, 80% forest

pop <- makeElephants(n, p, f)
x <- pop$x
z <- pop$z
```

```{r}
niter <- 100
res <- gibbs(x=x, k=2, niter=niter)
table(res$z[niter,],z)
res$Pi[niter,]
```

The sampler returns roughly similar proportions of each subpopulation $\Pi$,
and it does a poor job of differentiating between the two in terms of allele
frequencies and population of origin. This is understandable given that many
allele frequencies are not substantially different between forest and savannah
elephants at different loci, so some genotypes may have similar likelihoods for
both subpopulations.

If the Interpol allele frequencies are to be considered canonical and in no
need of estimation:

```{r}
niter <- 100
res <- gibbs(x=x, P=f, k=2, niter=niter)
table(res$z[niter,],z)
res$Pi[niter,]
```

The subpopulation frequencies returned are more consistent, though the
individual assignments remain poor.

When given larger differences between allele frequencies in the subpopulations,
the sampler better classifies individuals and returns more accurate estimates
allele frequencies.

```{r}
n <- 1000
fs <- c(.1, .1, .1, .1, .1)
ff <- c(.9, .9, .9, .9, .9)
f <- rbind(fs, ff)
p <- c(.2, .8) # 20% savannah, 80% forest

pop <- makeElephants(n, p, f)
x <- pop$x
z <- pop$z
```

```{r}
niter <- 100
res <- gibbs(x=x, k=2, niter=niter)
table(res$z[niter,],z)
res$Pi[niter,]
```

This returns better results for the proportion of savannah and forest elephants
and for the allele frequencies at each locus for each population.


**Note**, the sampler may return a 1 $\rightarrow$ 2 and 2 $\rightarrow$ 1
discrepancy. This is due to it being given uniform priors. Without an initial
bias toward some states being high or low frequency, the sampler cannot 'know'
beforehand which state value corresponds to what population.


# Extra credit

Extend your implementation to:

1. allow for an arbitrary $k$; and 
2. account for admixture.


## 1. Allow for an arbitrary $k$

The code above has already been extended to handle situations where k > 2.
Consider the following examples.

```{r}
n <- 1000
nloci <- 10
p <- c(.3, .2, .5)
f <- rbind(runif(nloci)/10, (runif(nloci)+5)/10, (runif(nloci)+9)/10)

pop <- makeElephants(n, p, f)
x <- pop$x
z <- pop$z
```

```{r}
niter <- 100
res <- gibbs(x=x, k=2, niter=niter)
table(res$z[niter,],pop$z)
res$Pi[niter,]
```

```{r}
niter <- 100
res <- gibbs(x=x, k=3, niter=niter)
table(res$z[niter,],pop$z)
res$Pi[niter,]
```

```{r}
niter <- 100
res <- gibbs(x=x, k=4, niter=niter)
table(res$z[niter,],pop$z)
res$Pi[niter,]
```

The sampler will return subpopulation frequencies and individual assignments
based on the number $k$ of clusters specified beforehand.


## 2. Account for admixture

Re-implement some functions to account for admixture. $Z$ must become a matrix
where each element corresponds to the origin at a particular locus.

```{r}
#' @param x an n by R matrix of data
#' @param P a K by R matrix of allele frequencies
#' @return an n by R matrix of group memberships for each allele
sample_z_admixture <- function(x, P, Pi){
  K <- nrow(P)
  loglik_matrix <- apply(x, 1, log_pr_x_given_P, P=P)

  lik_matrix <- exp(loglik_matrix)
  lik_matrix <- lik_matrix * as.vector(Pi)

  p.z.given.x <- apply(lik_matrix,2,normalize) # normalize columns

  z <- matrix(0, nrow(x), ncol(x))
  for(i in 1:nrow(z)) {
    for(j in 1:ncol(z)) {
      z[i,j] <- sample(1:K,size=1,prob=p.z.given.x[,i],replace=TRUE)
    }
  }
  return(z)
}

#' @param x an n by R matrix of data
#' @param z an n by R matrix of group memberships for each allele
#' @return a k by R matrix of allele frequencies
sample_P_admixture <- function(x,z,k){
  R <- ncol(x)
  P <- matrix(ncol=R,nrow=k)
  for(i in 1:k){
    sample_size <- colSums(apply(z == i, 2, as.numeric))
    number_of_ones <- colSums(x * apply(z == i, 2, as.numeric))
    P[i,] <- rbeta(R,1+number_of_ones,1+sample_size-number_of_ones) 
  }
  return(P)
}

#' @param z an n by R matrix of cluster allocations (1...k)
#' @param k the number of clusters
sample_pi_admixture <- function(z,k){
  counts <- matrix(0, nrow=1, ncol=k)
  for (i in 1:k) {
    counts[i] <- sum(z==i)
  }
  Pi <- gtools::rdirichlet(1,counts+1)
  return(Pi)
}

gibbs_admixture <- function(x, P=NULL, k = 2, niter = 100){
  z <- matrix(sample(1:k,length(x),replace=TRUE),nrow=nrow(x),ncol=ncol(x))
  Pi <- rep(1/k,k)

  res <- list(z = list(), Pi = list())
  res$z[[1]] <- z
  res$Pi[[1]] <- Pi
 
  pknown <- TRUE
  if (is.null(P)){
    pknown <- FALSE
  }

  for(i in 2:niter){
    Pi <- sample_pi_admixture(z,k)
    if (pknown == FALSE) {
      P <- sample_P_admixture(x,z,k)
    }
    z <- sample_z_admixture(x,P,Pi)
    res$z[[i]] <- z
    res$Pi[[i]] <- Pi
  }
  if (pknown == FALSE) {
    print(P)
  }
  return(res)
}
```


```{r}
# Generate the genotype and indicator for an elephant. For simplicity, only
# allow 1 recombination point.
makeAdmixedElephant <- function(z, f) {
  k <- length(z)   # number of subpopulations
  nloci <- ncol(f) # number of loci

  subpopulation1 <- sample(seq(1, k), 1, prob=z)
  subpopulation2 <- sample(seq(1, k), 1, prob=z)
  recombination <- sample(seq(2, nloci-2), 1) # don't allow recombinations at endpoints...

  elephant1 <- rbinom(f[subpopulation1,1:recombination], 1,
                      f[subpopulation1,1:recombination])
  elephant2 <- rbinom(f[subpopulation2,(recombination+1):nloci], 1,
                      f[subpopulation2,(recombination+1):nloci])
  
  elephant <- c(elephant1, elephant2)
  subpopulation <- c(rep(subpopulation1, recombination), rep(subpopulation2,
                                                             nloci-recombination))

  return(list(x=elephant, z=subpopulation))
}

# Generate genotypes and indicators for n elephants.
makeAdmixedElephants <- function(n, z, f) {
  genotypes <- matrix(0, n, ncol(f))
  indicators <- matrix(0, n, ncol(f))
  for (i in 1:n) {
    elephant <- makeAdmixedElephant(z, f)
    genotypes[i,] <- elephant$x
    indicators[i,] <- elephant$z
  }
  return(list(x=genotypes, z=indicators))
}
```

```{r}
n <- 1000
f1 <- c(.9, .9, .9, .9, .9, .9)
f2 <- c(.1, .1, .1, .1, .1, .1)
f <- rbind(f1, f2)
p <- c(.2, .8)

pop_mix <- makeAdmixedElephants(n, p, f)
x <- pop_mix$x
z <- pop_mix$z
```

```{r}
niter <- 100
res <- gibbs_admixture(x=x, k=2, niter=niter)
table(res$z[[niter]],z)
res$Pi[[niter]]
```

The Gibbs sampler when applied to samples with recent admixture (at most 1
recombination in each genotype) and **stark** differences in subpopulation
allele frequencies returns roughly the proportions of the original populations
and finds differences between allele frequencies in each.

The table at the bottom shows the proportion of correctly assigned loci. In
this example, 4850 loci are correctly identified and 1150 misidentified, and
85\% of forest loci and 70\% of savannah loci are correctly identified.

```{r}
sum(rowSums(z) == 12) + sum(rowSums(z) == 6) # number of genotypes without admixture
```

In the simulated data, 655 of 1000 individuals do not display any admixture,
so 2070 loci are found in interbred individuals.

Compare the effectiveness of the sampler for admixed versus non-admixed
individuals.

```{r}
z_adm_gibbs <- res$z[[niter]][rowSums(z) != 12 & rowSums(z) != 6,]
z_adm <- z[rowSums(z) != 12 & rowSums(z) != 6,]

z_noadm_gibbs <- res$z[[niter]][rowSums(z) == 12 | rowSums(z) == 6,]
z_noadm <- z[rowSums(z) == 12 | rowSums(z) == 6,]

table(z_adm_gibbs, z_adm)
table(z_noadm_gibbs, z_noadm)
```

For non-admixed individuals, the sampler correctly identifies 94\% of loci, and
for admixed individuals, the accuracy drops to 56\%. For the latter group, the
sampler isn't much better than a coin toss...

The result improves if not estimating the allele frequencies.

```{r}
niter <- 100
res <- gibbs_admixture(x=x, P=f, k=2, niter=niter)
table(res$z[[niter]],z)
res$Pi[[niter]]
```

```{r}
z_adm_gibbs <- res$z[[niter]][rowSums(z) != 12 & rowSums(z) != 6,]
z_adm <- z[rowSums(z) != 12 & rowSums(z) != 6,]

z_noadm_gibbs <- res$z[[niter]][rowSums(z) == 12 | rowSums(z) == 6,]
z_noadm <- z[rowSums(z) == 12 | rowSums(z) == 6,]

table(z_adm_gibbs, z_adm)
table(z_noadm_gibbs, z_noadm)
```

When $P$ is not being estimated, the assignments for the non-admixed
individuals improve, but those for admixed ones remain poor.

These results are for two starkly different subpopulations. The accuracy of the
Gibbs sampler is largely lost when considering similar subpopulations.
Consider the following example using admixed populations of forest and savannah
elephants.

```{r}
n <- 1000
fs <- c(0.40, 0.12, 0.21, 0.12, 0.02, 0.32)
ff <- c(0.80, 0.20, 0.11, 0.17, 0.23, 0.25)
f <- rbind(fs, ff)
p <- c(.2, .8) # 20% savannah, 80% forest

pop_mix <- makeAdmixedElephants(n, p, f)
x <- pop_mix$x
z <- pop_mix$z
```

```{r}
niter <- 100
res <- gibbs_admixture(x=x, k=2, niter=niter)
table(res$z[[niter]],z)
res$Pi[[niter]]
```

```{r}
z_adm_gibbs <- res$z[[niter]][rowSums(z) != 12 & rowSums(z) != 6,]
z_adm <- z[rowSums(z) != 12 & rowSums(z) != 6,]

z_noadm_gibbs <- res$z[[niter]][rowSums(z) == 12 | rowSums(z) == 6,]
z_noadm <- z[rowSums(z) == 12 | rowSums(z) == 6,]

table(z_adm_gibbs, z_adm)
table(z_noadm_gibbs, z_noadm)
```

The estimates for $P$, $\Pi$, and $Z$ are largely incorrect.

The results improve if not estimating the allele frequencies.

```{r}
niter <- 100
res <- gibbs_admixture(x=x, P=f, k=2, niter=niter)
table(res$z[[niter]],z)
res$Pi[[niter]]
```

```{r}
z_adm_gibbs <- res$z[[niter]][rowSums(z) != 12 & rowSums(z) != 6,]
z_adm <- z[rowSums(z) != 12 & rowSums(z) != 6,]

z_noadm_gibbs <- res$z[[niter]][rowSums(z) == 12 | rowSums(z) == 6,]
z_noadm <- z[rowSums(z) == 12 | rowSums(z) == 6,]

table(z_adm_gibbs, z_adm)
table(z_noadm_gibbs, z_noadm)
```

Here, the correct value for $\Pi$ is returned, and the assignments for
non-admixed individuals are much better. The ones for admixed individuals are
still poor, essentially a coin toss.
