---
title: Homework 8
author: Ansel George
output: pdf_document
---


```{r, message=F}
library(ggplot2)
library(dplyr)
library(tidyr)
library(ape)

set.seed(10)
```


# 1 Cancer cell population model

*Background: Last week in Problem 1.2, we used a deterministic model of
exponential growth for a cancer cell population.*

> _Problem 1.2. Consider modeling cancer cells entering the blood/circulatory
> systems as a non-homogeneous Poisson Process. In particular, suppose there is
> a source cancer cell population whose size is growing deterministically. More
> specifically, assume $x(t)$, the size of the source population at time $t$,
> is growing exponentially at rate $r$ per hour (i.e. $x(t) = x_0 e^{rt}$ )
> where $x_0$ is the starting size of the population. An expected fraction $m$
> (for $0 \leq m \leq 1$) of that cancer population is entering the
> blood/circulatory system, such that cancer cells enter the blood as a Poisson
> Process with an instantaneous rate of $mx(t)$ per hour._

*Let's instead consider a model of a cancer cell population size, $N(t)$, as a
pure birth (Yule) process in which each cell gives birth at rate $r$ per hour.
That is, the general birth process where $\lambda_i = ri$ and $\mu_i = 0$ for
all $i$. Suppose the cancer population starts with ten cells and assume $r =
0.01$.*

*Answer the following questions:*

1. *Assuming $r = 0.01$, after how many hours should we expect an $e$-fold
   change in the mean value?*

The probability of transitioning from state $i$ to state $j$ is
negative-binomially distributed.

\begin{align}
\mathbb{P}_{ij(t)} &= { j-1 \choose i-1 } (e^{-rt})^i (1-e^{-rt})^{j-i}
\end{align}

This distribution has expected value:

\begin{align}
\mathbb{E}[\mathbb{P}_{ij}(t)] &= \frac{i}{p} \\
  &= \frac{i}{e^{-rt}}
\end{align}

The expected time needed to elapse before fold-change $k = \frac{j}{i}$ is as
follows:

\begin{align}
\mathbb{E}[\mathbb{P}_{ij}(t_0)] &= \frac{i}{e^{-r0}} \\
  &= i \\
\mathbb{E}[\mathbb{P}_{ij}(t)] &= \frac{i}{e^{-rt}} \\
\end{align}

\begin{align}
\textrm{When }\mathbb{E}[\mathbb{P}_{ij}(t)] &= k\mathbb{E}[\mathbb{P}_{ij}(t_0)], \\
\frac{i}{e^{-rt}} &= k i \\
  &= \frac{j}{i} i \\
  &= j \\
\implies \frac{1}{e^{-rt}} &= \frac{j}{i} \\
\implies e^{rt} &= \frac{j}{i} \\
\implies t &= \frac{1}{r} ln\big(\frac{j}{i}\big)
\end{align}

*To illustrate the above relationship, consider the following simulation:

```{r}
# Simulate a Markov chain starting with `n0` cells and growth rate `r` that
# grows for `steps` steps.
simulateCells <- function(n0, steps, r) {
  sizehistory <- rep(0, steps)
  sizehistory[1] <- n0
  timehistory <- rep(0, steps)
  for (i in 2:steps) {
    timehistory[i] <- timehistory[i-1] + rexp(1, r*sizehistory[i-1])
    sizehistory[i] <- sizehistory[i-1] + 1
  }
  return(list(times=timehistory,popsize=sizehistory))
}

# Compute the expected time to fold change based on above analytical result.
foldExpectationTime <- function(t, r, n0) {
  return(1/r*log(j/n0))
}
```

```{r}
n0 <- 10      # initial number of cells
steps <- 500  # number of steps in the embedded chain
reps <- 10000 # number of replicates
r <- .01      # cell growth rate

# Run many replicates of the above simulation function
popsizes <- matrix(0, nrow=reps, ncol=steps)
times <- matrix(0, nrow=reps, ncol=steps)

for (i in 1:reps) {
  res <- simulateCells(n0, steps, r)
  popsizes[i,]  <- res$popsize
  times[i,]  <- res$times
}
```

```{r}
j <- apply(popsizes, 2, mean)    # population sizes from simulation
k <- apply(popsizes, 2, mean)/n0 # fold change at each step in Markov chain
t_mean <- apply(times, 2, mean)  # average time to reach population size

# plot(k, t_mean)
plot(j, t_mean, xlab='j', ylab='time (hr)')
lines(j, foldExpectationTime(j, r, n0), col='red')

qqplot(t_mean, foldExpectationTime(j, r, n0), xlab='sim. time (hr)', ylab='exp time (hr)')
abline(0,1,col='red')

```

The following are some checks to ensure that the simulation is correct by
testing whether it has the properties of the analytical distribution.

1. Check that when starting at $n_0 > 10$, the population still follows the
   same expectation for fold-change.

```{r}
offset_t <- 50
k2 <- apply(popsizes[,(offset_t+1):ncol(popsizes)], 2, mean)/(n0+offset_t)
j2 <- apply(popsizes[,(offset_t+1):ncol(popsizes)], 2, mean)
t_mean2 <- apply(times[,(offset_t+1):ncol(popsizes)], 2, mean)-mean(times[,offset_t+1])

plot(k, t_mean, xlab='fold-change k', ylab='time (hr)')
points(k2, t_mean2, col='blue')
```

2. Check that the estimate for the population growth rate $\frac{1}{\bar{X}}$
   at each step matches that for an exponential distribution, which here is
   $rj$.

```{r}
t_diff <- t_mean[2:length(t_mean)]-t_mean[1:(length(t_mean)-1)]
plot(j[2:length(j)], t_diff, xlab='j', ylab='delta time (hr)')
lines(j, 1/r/j, col='red')

plot(j[2:length(j)], 1/r/t_diff, xlab='j', ylab='delta time (hr)')
abline(0,1,col='red')
```


2. _Assuming $r = 0.01$, use known exact solutions for $\mathbb{P}\big(N (t) =
   j\big)$ to plot the value of $\mathbb{P}\big(N (t) = j\big)$ for $1 \leq j
   \leq 100$ for four values of $t$ (in hours): $t \in \{0, 50, 100, 200\}$. Do
   this using a vertically stacked multi-panel plot. (Hint: If you use the
   negative binomial function in R be conscious of the parameterization!)_

```{r}
probnbinom <- function(i,j,t,r) {
  if (t==0) {
    # return 1 when j=i at time 0
    return((choose(j-1,i-1)*((exp(-r*t))^i)*(1-exp(-r*t))^(j-i))*(j==i)) 
  } else {
    return((choose(j-1,i-1)*((exp(-r*t))^i)*(1-exp(-r*t))^(j-i))*(j>i))
  }
}
```

```{r}
i <- 10
j <- seq(i,100,.1)
r <- .01

fr <- tbl_df(data.frame(j=j, T000=probnbinom(n0,j,0,r),
                        T050=probnbinom(n0,j,50,r),
                        T100=probnbinom(n0,j,100,r),
                        T200=probnbinom(n0,j,200,r)))
fr %>% gather(Time,p,-j) %>%
  ggplot() + aes(x=j, y=p) + facet_grid(Time~.) + geom_line()
```

To make later distributions more visible, below is the same plot with y-axis
scales free.

```{r}
fr %>% gather(Time,p,-j) %>%
  ggplot() + aes(x=j, y=p) + facet_grid(Time~.,scale='free') + geom_line()
```


3. _What is the mean and variance of the cancer cell population size after $200$
   hours?_

This distribution has expected value:

\begin{align}
\frac{i}{p} &= \frac{i}{e^{-rt}} \\
  &= \frac{10}{e^{-rt}} \\
  &= \frac{10}{e^{-.01 * 200}} \\
  &= 73.89056
\end{align}

And variance:

\begin{align}
\frac{i(1-p)}{p^2} &= \frac{i(1-e^{-rt})}{e^{-2rt}} \\
  &= i\big(e^{2rt}-e^{rt}\big)\\
  &= 10\big(e^{2 * .01 * 200} - e^{.01 * 200}\big)\\
  &= 472.0909
\end{align}


4. _**Extra Credit**. Revisit last week’s problem in which a proportion $m =
   0.0001$ of cancer cells enter the blood stream, and we asked at what time is
   there a $99%$ probability that the blood will have received $1000$ cancer
   cells. Update your answer using our new stochastic model for the cancer cell
   source population size. (Continue to assume $r = 0.01$ and the starting
   population is size in the cancer is $10$ cells). Provide a value to the
   nearest tenth of an hour and assess if the number is higher or lower than in
   the deterministic model._

Under the deterministic model, without cell death in the bloodstream, it took
$\sim 928.4$ hours for there to be at least 1000 cells in the bloodstream with
99\% probability. With death, that time extended to 997.6803 hours.

Under a stochastic model:

With each step in the stochastic model, there are $n$ cells in the population,
and at each step $m * n * \delta t$ cells enter the bloodstream, where $m$ is
the entry rate and $\delta t$ is the instantaneous time differential.
(Non-infinitesimal values for $\delta t$ as done in the below simulation will
introduce error.)

The total number of cells in the bloodstream $x$ is the sum of $m n_i$ where
$n_i$ is the number of cells in the cancer population at the $i$th step in the
embedded chain.

For $x=1000$, the number of steps in the chain needed to reach that threshold
is:

```{r}
simulateBloodCells <- function(n0, r, m, reps, u=0) {
  elapsed <- rep(0, reps)
  bloodcells <- rep(0, reps)
  population <- rep(n0, reps)

  for (i in 1:reps) {
    while(bloodcells[i]<1000) {
      delta_t <- rexp(1, r*population[i])
      elapsed[i] <- elapsed[i] + delta_t
      population[i] <- population[i] + 1
      bloodcells[i] <- bloodcells[i] + m*population[i]*delta_t - u*(bloodcells[i])*delta_t
    }
  }
  return(list(times=elapsed,popsize=population,bloodsize=bloodcells))
}
```

```{r}
r <- .01
m <- .0001
n0 <- 10
u <- .01

reps <- 100 # very time-consuming; should parallelize but can't be bothered...

res_nodeath <- simulateBloodCells(n0, r, m, reps)
res_death <- simulateBloodCells(n0, r, m, reps, u)
```

```{r}
hist(res_nodeath$times, xlab='time (hr)')
mean(res_nodeath$times)
var(res_nodeath$times)
quantile(res_nodeath$times, .99)

hist(res_death$times, xlab='time (hr)')
mean(res_death$times)
var(res_death$times)
quantile(res_death$times, .99)
```

Using the 99\% quantile for the distribution of times when $x=1000$ is reached,
it takes 1013.4 hours to cross the threshold of 1000 cells in the bloodstream.
The waiting time is 1045.4 hours when cells in the bloodstream are dying. Both
numbers are larger than the expected times from the deterministic models.

Note that estimates are skewed toward larger values due to large timesteps
taken especially near the beginning of the simulation and due to only
estimating waiting time from values of the blood population that exceed the
threshold $x=1000$.


# 2 Basic molecular phylogenetics

_Background: Consider a three taxon tree of human, chimp, and gorilla. Assume
the human-chimp species divergence is $6$ million years ago. Assume the
human/chimp vs. gorilla divergence is $7$ million years ago. Assume a
substitution rate of $\mu = 0.2 * 10^{-8}$ per base pair (bp) per year. Further
assume the **Jukes-Cantor** model of DNA evolution and that the initial bp in
the common ancestor of all three species is equally likely to have been an $A$,
$C$, $T$, or $G$. The Jukes-Cantor model assumes that mutations of all types
are equally likely. (Note: Ignore the subtlety that sequence divergence times
often predate species divergence times)._

*Answer the following questions:*

1. *Write the rate matrix ($Q$) and the transition function ($P_{ij}(t)$) for
   the Jukes-Cantor model.*

\[ Q = \left[
  \begin{array}{cccc}
    -\frac{3}{4}\mu & \frac{1}{4}\mu  & \frac{1}{4}\mu  & \frac{1}{4}\mu \\
    \frac{1}{4}\mu  & -\frac{3}{4}\mu & \frac{1}{4}\mu  & \frac{1}{4}\mu \\
    \frac{1}{4}\mu  & \frac{1}{4}\mu  & -\frac{3}{4}\mu & \frac{1}{4}\mu \\
    \frac{1}{4}\mu  & \frac{1}{4}\mu  & \frac{1}{4}\mu  & -\frac{3}{4}\mu \\
  \end{array}
\right] \]

\begin{align}
P_{ij, j \neq i}(t) &= \frac{1}{4} - \frac{1}{4}e^{-\mu t} \\
P_{ii}(t) &= \frac{1}{4} + \frac{3}{4}e^{-\mu t}
\end{align}

\[ P(t) = \frac{1}{4} \left[ 
  \begin{array}{cccc}
    1 & 1 & 1 & 1 \\
    1 & 1 & 1 & 1 \\
    1 & 1 & 1 & 1 \\
    1 & 1 & 1 & 1 \\
  \end{array}
\right] + \frac{1}{4} \left[
  \begin{array}{cccc}
    3  & -1 & -1 & -1 \\
    -1 & 3  & -1 & -1 \\
    -1 & -1 & 3  & -1 \\
    -1 & -1 & -1 & 3 \\
  \end{array}
\right] e^{-\mu t}
\]



2. *Compute the log-likelihood of a site configuration where: human=$A$,
   chimp=$A$, gorilla=$G$ under the given phylogenetic tree. Show your
   calculations and/or code.*

```{r}
tree <- read.tree(text="((Human=A,Chimp=A),Gorilla=G);")
plot(tree, type='cladogram')
nodelabels()
```

\begin{align}
P(tree) &= \sum_{z_4} \sum_{z_5} P(Gorilla=G|Z_4) P(Chimp=A|Z_5) P(Human=A|Z_5) P(Z_5|Z_4) P(Z_5) \\
  &= \sum_{z_4} P(Gorilla=G|Z_4) P(Z_4) \sum_{z_5} P(Chimp=A|Z_5) P(Human=A|Z_5) P(Z_5|Z_4) \\
  &= \sum_{z_4} P(G|Z_4) P(Z_4) \sum_{z_5} P(A|Z_5) P(A|Z_5) P(Z_5|Z_4) \\
  &= \sum_{z_4} P(G|Z_4) P(Z_4) m_5(z_4)
\end{align}

where

\begin{align}
 m_5(z_4) &= \sum_{z_5} P(A|Z_5) P(A|Z_5) P(Z_5|Z_4) \\
  &= P(A|Z_4)P(A|A)^2 + P(C|Z_4)P(A|C)^2 + P(G|Z_4)P(A|G)^2 + P(T|Z_4)P(A|T)^2
\end{align}

Therefore,

\begin{align}
P(tree) &= \sum_{z_4} P(G|Z_4) P(Z_4) m_5(z_4) \\
  &= P(G|A) P(A) m_5(A) + P(G|C) P(C) m_5(C) + P(G|G) P(G) m_5(G) + P(G|T) P(T) m_5(T)
\end{align}


In the following code, the dummy variables $\{1,2,3,4\}$ correspond to
$\{A,C,G,T\}$. Also, the probability for $P(Z_4)$ is assumed to be drawn from
the stationary distribution for the Jukes-Cantor model, which is $\frac{1}{4}$
for each base. Because the likelihoods are equal, they do not contribute to
differences in the log-likelihoods for any given tree marginalized over hidden
states, so they are ignored in the following computations.

```{r}
m5_hc <- function(z4, Pt) {
  return(Pt[z4,] %*% as.matrix(nrow=4, ncol=1, Pt_hg[,1]^2))
}
```

```{r}
mu <- .2 * 10^(-8)

t_hc <- 6*10^6
t_hg <- 7*10^6

Pt_hc <- 1/4*(matrix(1, nrow=4, ncol=4) +
              (matrix(-1, nrow=4, ncol=4) + 4*diag(4))*exp(-mu*t_hc))
Pt_hg <- 1/4*(matrix(1, nrow=4, ncol=4) +
              (matrix(-1, nrow=4, ncol=4) + 4*diag(4))*exp(-mu*t_hg))

lik <- Pt_hg[,3] %*% m5_hc(c(1,2,3,4), Pt_hc)
loglik <- log(lik)
loglik
```


3. _Compute the log-likelihood for the same site configuration under a revised
   phylogeny where humans and gorillas now separate $6$ million years ago and
   chimpanzees separate from the human/gorilla common ancestor $7$ million
   years ago._

```{r}
tree <- read.tree(text="((Human=A,Gorilla=G),Chimp=A);")
plot(tree, type='cladogram')
nodelabels()
```

\begin{align}
P(tree) &= \sum_{z_4} \sum_{z_5} P(Gorilla=G|Z_4) P(Chimp=A|Z_5) P(Human=A|Z_5) P(Z_5|Z_4) P(Z_5) \\
  &= \sum_{z_4} P(Z_4) P(Chimp=A|Z_4) \sum_{z_5} P(Gorilla=G|Z_5) P(Human=A|Z_5) P(Z_5|Z_4) \\
  &= \sum_{z_4} P(Z_4) P(A|Z_4) \sum_{z_5} P(G|Z_5) P(A|Z_5) P(Z_5|Z_4) \\
  &= \sum_{z_4} P(G|Z_4) P(Z_4) m_5(z_4)
\end{align}

where

\begin{align}
 m_5(z_4) &= \sum_{z_5} P(A|Z_5) P(A|Z_5) P(Z_5|Z_4) \\
  &= P(A|Z_4)P(A|A)P(G|A) + P(C|Z_4)P(A|C)P(G|C) + P(G|Z_4)P(A|G)P(G|G) + P(T|Z_4)P(A|T)P(G|T)
\end{align}

Therefore,

\begin{align}
P(tree) &= \sum_{z_4} P(A|Z_4) P(Z_4) m_5(z_4) \\
  &= P(A|A) P(A) m_5(A) + P(A|C) P(C) m_5(C) + P(A|G) P(G) m_5(G) + P(A|T) P(T) m_5(T)
\end{align}

```{r}
m5_hg <- function(z4, Pt) {
  return(Pt[z4,] %*% as.matrix(nrow=4, ncol=1, (Pt_hg[,3]*Pt_hg[,1])))
}
```

```{r}
mu <- .2 * 10^(-8)

t_hc <- 7*10^6
t_hg <- 6*10^6

Pt_hc <- 1/4*(matrix(1, nrow=4, ncol=4) +
              (matrix(-1, nrow=4, ncol=4) + 4*diag(4))*exp(-mu*t_hc))
Pt_hg <- 1/4*(matrix(1, nrow=4, ncol=4) +
              (matrix(-1, nrow=4, ncol=4) + 4*diag(4))*exp(-mu*t_hg))

lik2 <-  Pt_hc[,1] %*% (m5_hg(c(1,2,3,4), Pt_hg))
loglik2 <- log(lik2)
loglik2
```

4. _Compare your results for 2 and 3 - which tree is favored and by how many
   log-likelihood units?  What is your judgment on whether there is enough
   support to favor one tree versus another. (Put another way, would you want
   to collect more data or can you already reach a confident decision on the
   phylogeny?)_


```{r}
loglik-loglik2
exp(loglik-loglik2)
```

The tree where humans and chimpanzees share the most recent common ancestor is
more likely than the one where humans and gorillas are more recently derived.
The elevated likelihood is around $\sim 0.77$ log-likelihood units (a $\sim
2.16$-fold likelihood-ratio).

This locus alone, though, is not sufficient to establish with confidence that
chimpanzees and humans have most recently diverged. A likelihood-ratio test
(e.g $\chi^2$) can be used to more formally assess significance.

```{r}
1-pchisq(loglik-loglik2, 2)
```


5. _Research the **Kimura two-parameter** and the **General Time Reversible**
   models of sequence evolution. Both of these models are more frequently used
   than the Jukes-Cantor model. Describe in one sentence what is useful about
   each model._

**Kimura two-parameter model:** Allows for different probabilities of
transitions and transversions, but keeps the uniform stationary distribution.

**General Time Reversible model:** Allows for non-uniform stationary
distributions of nucleotide frequencies and establishes time-reversibility of
nucleotide substitutions.
